
---
title: "Baseflow Data Processing Report (1945-2012 Historical + Future RCPs)"
subtitle: "Temporal Aggregation and Spatial Resolution Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

## 1. Temporal Drift: Is the Region Getting Wetter?

First, we analyze if the maximum baseflow intensity has increased over the observed period and how it is projected to change under different emission pathways.

```{r setup, include=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(viridis)

# Load the processed yearly data
yearly_max_hist <- rast("yearly_max_1945_2012.tif")
monthly_max_hist <- rast("monthly_max_1945_2012.tif")

yearly_max_rcp45 <- rast("yearly_max_RCP45_1945_2099.tif")
monthly_max_rcp45<- rast("monthly_max_RCP45_1945_2099.tif")

yearly_max_rcp85 <- rast("yearly_max_RCP85_1945_2099.tif")
monthly_max_rcp85<- rast("monthly_max_RCP85_1945_2099.tif")


# Define file paths with your updated filenames
file_history <- "netCDF_data/Gridded 1945-2012 Baseflow.nc"
files_rcp45 <- c(
  "netCDF_data/CanESM2 RCP45 Baseflow 1945-2099 1.nc",
  "netCDF_data/CanESM2 RCP45 Baseflow 1945-2099 3.nc", 
  "netCDF_data/CanESM2 RCP45 Baseflow 1945-2099 2.nc"
)
files_rcp85 <- c(
  "netCDF_data/CanESM2 RCP85 Baseflow 1945-2099 1.nc",
  "netCDF_data/CanESM2 RCP85 Baseflow 1945-2099 3.nc",
  "netCDF_data/CanESM2 RCP85 Baseflow 1945-2099 2.nc"
)


# Load the Daily NetCDF files (terra will stitch the vectors together)
#take a bit of time
bf_stack_hist <- rast(file_history)["BASEFLOW"]
bf_stack_rcp45 <- rast(files_rcp45)["BASEFLOW"]
bf_stack_rcp85 <- rast(files_rcp85)["BASEFLOW"]

  

num_layers <- nlyr(yearly_max_rcp45)
# 2. Create a sequence of years
start_year <- 1945
years_seq <- start_year : (start_year + num_layers - 1)
# 3. Assign the time attribute
time(yearly_max_hist) <- as.Date(paste0(1945:2012, "-01-01"))
time(yearly_max_rcp45) <- as.Date(paste0(years_seq, "-01-01"))
time(yearly_max_rcp85) <- as.Date(paste0(years_seq, "-01-01"))

# 1. Calculate the regional average of the yearly maximums
regional_annual_max_hist <- global(yearly_max_hist, "mean", na.rm = TRUE)
regional_annual_max_rcp45 <- global(yearly_max_rcp45, "mean", na.rm = TRUE)
regional_annual_max_rcp85 <- global(yearly_max_rcp85, "mean", na.rm = TRUE)

# 2. Create dataframes for plotting
df_historic <- data.frame(Year = 1945:2012, Max_Baseflow = regional_annual_max_hist$mean, Scenario = "Historical")
df_rcp45 <- data.frame(Year = 1945:2099, Max_Baseflow = regional_annual_max_rcp45$mean, Scenario = "RCP 4.5")
df_rcp85 <- data.frame(Year = 1945:2099, Max_Baseflow = regional_annual_max_rcp85$mean, Scenario = "RCP 8.5")

# Combine all three
df_trend <- bind_rows(df_historic, df_rcp45, df_rcp85)

#check the first few rows for 3 scenarios
#use filter
head(df_trend %>% filter(Scenario == "Historical"))
head(df_trend %>% filter(Scenario == "RCP 4.5"))
head(df_trend %>% filter(Scenario == "RCP 8.5"))


```

```{r 1.5}
# 3. Plot with a trend line (Loess smoothing)
ggplot(df_trend, aes(x = Year, y = Max_Baseflow, color = Scenario, fill = Scenario)) +
  geom_line(alpha = 0.5) +
  geom_point(alpha = 0.7) +
  #geom_smooth(method = "loess") +
  scale_color_manual(values = c("Historical" = "black", "RCP 4.5" = "steelblue", "RCP 8.5" = "darkred")) +
  scale_fill_manual(values = c("Historical" = "grey", "RCP 4.5" = "lightblue", "RCP 8.5" = "pink")) +
  theme_minimal() +
  labs(title = "Average Annual Maximum Baseflow (1945-2099)",
       subtitle = "Aggregated across the Region",
       y = "Baseflow (mm/day)",
       x = "Year")

#seperate the plot into three plots for each scenario
ggplot(df_trend, aes(x = Year, y = Max_Baseflow, color = Scenario, fill = Scenario)) +
  geom_line(alpha = 0.5) +
  geom_point() +
  #geom_smooth(method = "loess") +
  scale_color_manual(values = c("Historical" = "black", "RCP 4.5" = "steelblue", "RCP 8.5" = "darkred")) +
  scale_fill_manual(values = c("Historical" = "grey", "RCP 4.5" = "lightblue", "RCP 8.5" = "pink")) +
  theme_minimal() +
  labs(title = "Average Annual Maximum Baseflow (1945-2099)",
       subtitle = "Aggregated across the Vancouver/PNW Region (Including RCP Scenarios)",
       y = "Baseflow (mm/day)",
       x = "Year") +
  facet_wrap(~ Scenario, ncol = 1)  

```

---

## 2. Spatial Visualization (Snapshots)

Let’s visualize the hydrology data for specific dates to ensure the spatial patterns look correct across all three datasets.

```{r 2}
# === HISTORICAL SNAPSHOTS ===
target_dates_hist <- as.Date(c("1945-01-01", "1972-05-01", "1972-07-01", "1997-06-01"))
indices_hist <- which(as.Date(time(bf_stack_hist)) %in% target_dates_hist)

if (length(indices_hist) > 0) {
  par(mfrow = c(2, 2))
  for (i in indices_hist) {
    plot(bf_stack_hist[[i]], main = paste("Historic:", time(bf_stack_hist)[i]), col = topo.colors(50), axes = FALSE) 
  }
}

# === RCP 4.5 SNAPSHOTS ===
target_dates_future <- as.Date(c("2030-01-01", "2050-05-01", "2070-07-01", "2099-06-01"))
indices_rcp45 <- which(as.Date(time(bf_stack_rcp45)) %in% target_dates_future)

if (length(indices_rcp45) > 0) {
  par(mfrow = c(2, 2))
  for (i in indices_rcp45) {
    plot(bf_stack_rcp45[[i]], main = paste("RCP 4.5:", time(bf_stack_rcp45)[i]), col = topo.colors(50), axes = FALSE) 
  }
}

# === RCP 8.5 SNAPSHOTS ===
indices_rcp85 <- which(as.Date(time(bf_stack_rcp85)) %in% target_dates_future)

if (length(indices_rcp85) > 0) {
  par(mfrow = c(2, 2))
  for (i in indices_rcp85) {
    plot(bf_stack_rcp85[[i]], main = paste("RCP 8.5:", time(bf_stack_rcp85)[i]), col = topo.colors(50), axes = FALSE) 
  }
}
par(mfrow = c(1, 1)) # Reset

```

---

## 3. Interactive Maps for Key Dates, a decent visualization that can compare the spatial patterns directly for the same time slice across the three dataset, or different time slices within the same dataset.
```{r 3}
library(leaflet)
library(terra)
library(htmltools)

# 1. Define your list of target dates
target_dates <- as.Date(c("2012-06-08", "2026-03-08", "2070-01-01"))

# 2. Create a list to store the maps
map_list <- lapply(target_dates, function(date_val) {
  
  # Extract layers for the specific date
  get_layer <- function(stack, d) {
    idx <- which(time(stack) == d)
    if(length(idx) > 0) return(stack[[idx]]) else return(NULL)
  }
  
  l_hist <- get_layer(bf_stack_hist, date_val)
  l_45   <- get_layer(bf_stack_rcp45, date_val)
  l_85   <- get_layer(bf_stack_rcp85, date_val)
  
  # Identify available groups
  available_groups <- c()
  valid_layers <- list()
  
  if(!is.null(l_hist)) {
    available_groups <- c(available_groups, "Historical")
    valid_layers$Historical <- l_hist
  }
  if(!is.null(l_45)) {
    available_groups <- c(available_groups, "RCP 4.5")
    valid_layers$RCP45 <- l_45
  }
  if(!is.null(l_85)) {
    available_groups <- c(available_groups, "RCP 8.5")
    valid_layers$RCP85 <- l_85
  }
  
  # Determine max value for this specific date for color scaling
  # (Ensures all buttons for THIS map share the same color logic)
  max_val <- max(sapply(valid_layers, function(x) max(values(x), na.rm=TRUE)), na.rm=TRUE)
  
  pal_shared <- colorNumeric(
    palette = c("#FFFFCC", "#41B6C4", "#0C2C84"), 
    domain = c(0, max_val),
    na.color = "transparent"
  )
  
  # Build the Map
  m <- leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    addControl(html = paste("<strong>Map Date:", date_val, "</strong>"), position = "topright")
  
  if(!is.null(l_hist)) m <- m %>% addRasterImage(l_hist, colors = pal_shared, opacity = 0.8, group = "Historical", project = FALSE)
  if(!is.null(l_45))   m <- m %>% addRasterImage(l_45, colors = pal_shared, opacity = 0.8, group = "RCP 4.5", project = FALSE)
  if(!is.null(l_85))   m <- m %>% addRasterImage(l_85, colors = pal_shared, opacity = 0.8, group = "RCP 8.5", project = FALSE)
  
  m <- m %>%
    addLegend(pal = pal_shared, values = c(0, max_val), title = "Baseflow (mm)", position = "bottomright") %>%
    addLayersControl(
      baseGroups = available_groups,
      options = layersControlOptions(collapsed = FALSE)
    )
  
  return(m)
})

# 3. Render all maps in the R Markdown output
tagList(map_list)
```

---

## 4. Point Extraction Daily

We extract a time series for a specific coordinate across all three datasets to compare the daily volatility.

```{r 4}
# 1. Find valid land coordinates
first_layer <- yearly_max_hist[[1]]
valid_points <- as.points(first_layer, values=TRUE, na.rm=TRUE)

# 2. Pick a safe sample point
idx <- min(1000, length(valid_points))
sample_point <- valid_points[idx]
coords <- crds(sample_point)

print(paste("Extracting data for Lat:", round(coords[1,2], 2), "Lon:", round(coords[1,1], 2)))

# 3. Extract time series for this location across all stacks
point_vals_hist <- as.numeric(terra::extract(bf_stack_hist, sample_point, ID=FALSE))
point_vals_rcp45 <- as.numeric(terra::extract(bf_stack_rcp45, sample_point, ID=FALSE))
point_vals_rcp85 <- as.numeric(terra::extract(bf_stack_rcp85, sample_point, ID=FALSE))

# 4. Create dataframes and combine
ts_hist <- data.frame(Date = time(bf_stack_hist), Baseflow = point_vals_hist, Scenario = "Historical")
ts_rcp45 <- data.frame(Date = time(bf_stack_rcp45), Baseflow = point_vals_rcp45, Scenario = "RCP 4.5")
ts_rcp85 <- data.frame(Date = time(bf_stack_rcp85), Baseflow = point_vals_rcp85, Scenario = "RCP 8.5")

point_ts_all <- bind_rows(ts_hist, ts_rcp45, ts_rcp85)

# Filter out early anomalies if necessary (like 1945-01-01)
point_ts_all <- point_ts_all %>% filter(Date >= as.Date("1945-02-02"))

# 5. Plot Interactive Daily Series
library(plotly)
p_daily <- ggplot(point_ts_all, aes(x = Date, y = Baseflow, color = Scenario)) +
  geom_line(alpha = 0.6, linewidth = 0.5) +
  scale_color_manual(values = c("Historical" = "darkgreen", "RCP 4.5" = "steelblue", "RCP 8.5" = "darkred")) +
  theme_light() +
  labs(title = paste("Local Daily Baseflow Series (1945-2099)"), 
       subtitle = paste("Coordinates:", round(coords[1,1], 2), ",", round(coords[1,2], 2)),
       y = "Baseflow (mm)")

ggplotly(p_daily)

#seperate the plot into three plots for each scenario
p_daily_facet <- ggplot(point_ts_all, aes(x = Date, y = Baseflow, color = Scenario)) +
  geom_line(alpha = 0.6, linewidth = 0.5) +
  scale_color_manual(values = c("Historical" = "darkgreen", "RCP 4.5" = "steelblue", "RCP 8.5" = "darkred")) +
  theme_light() +
  labs(title = paste("Local Daily Baseflow Series (1945-2099)"), 
       subtitle = paste("Coordinates:", round(coords[1,1], 2), ",", round(coords[1,2], 2)),
       y = "Baseflow (mm)") +
  facet_wrap(~ Scenario, ncol = 1)

ggplotly(p_daily_facet)

```
---

## 5. Identifying the "Record Breaker" Year

Which year actually held the highest baseflow pixel-wise for each scenario?

```{r 5}
# Find the index (year) of the maximum value for every pixel
year_of_max_hist <- which.max(yearly_max_hist)
historic_max_years <- year_of_max_hist + 1944

year_of_max_45 <- which.max(yearly_max_rcp45)
future_max_years_45 <- year_of_max_45 + 1944

year_of_max_85 <- which.max(yearly_max_rcp85)
future_max_years_85 <- year_of_max_85 + 1944

par(mfrow = c(1, 3))
plot(historic_max_years, main = "Historic Max Year", col = terrain.colors(68), axes = FALSE)
plot(future_max_years_45, main = "RCP 4.5 Max Year", col = heat.colors(155), axes = FALSE)
plot(future_max_years_85, main = "RCP 8.5 Max Year", col = heat.colors(155), axes = FALSE)
par(mfrow = c(1, 1))

```

```{r 5.5}
library(terra)
library(leaflet)

# Bin years into numeric decades
decades_rast_hist <- floor(historic_max_years / 10) * 10
decades_rast_45 <- floor(future_max_years_45 / 10) * 10
decades_rast_85 <- floor(future_max_years_85 / 10) * 10

# Create Color Palettes
pal_hist <- colorNumeric(palette = c("#D73027", "#4575B4"), domain = range(values(decades_rast_hist), na.rm=TRUE), na.color = "transparent")
pal_future <- colorNumeric(palette = "YlOrRd", domain = range(values(decades_rast_85), na.rm=TRUE), na.color = "transparent")

# Create the Interactive Map with layers for all three
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addRasterImage(decades_rast_hist, colors = pal_hist, opacity = 0.8, group = "Historical Era") %>%
  addRasterImage(decades_rast_45, colors = pal_future, opacity = 0.8, group = "RCP 4.5 Era") %>%
  addRasterImage(decades_rast_85, colors = pal_future, opacity = 0.8, group = "RCP 8.5 Era") %>%
  
  addLegend(pal = pal_hist, values = values(decades_rast_hist), title = "Historic Era", position = "bottomright") %>%
  addLegend(pal = pal_future, values = values(decades_rast_85), title = "Future Era (RCPs)", position = "bottomleft") %>%
  
  addLayersControl(
    baseGroups = c("Historical Era", "RCP 4.5 Era", "RCP 8.5 Era"),
    options = layersControlOptions(collapsed = FALSE)
  )

```

---

## 6. Statistical Distribution (Ridge Plot)

Since you have the `monthly_max_1945_2012.tif`, you can visualize how the distribution of "maximums" shifts through the seasons.

---

## Summary of Findings to look for:

1. **Topography Influence:** You should notice the highest values concentrated along the Coast Mountains and the Cascades, where orographic lift intensifies rainfall.
2. **The 1990s/2000s Shift:** Check Section 1's trend line; many PNW models show a slight intensification of extreme events in the latter half of the 20th century.
3. **Pixel Size:** Remember that your pixels are large enough to encompass entire small sub-watersheds, meaning these baseflow values represent a significant volume of water moving into the stream network.

```

