---
title: "Baseflow Data Processing Report (1945-2012)"
subtitle: "Temporal Aggregation and Spatial Resolution Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

## 1. Temporal Drift: Is the Region Getting Wetter?

First, we analyze if the maximum baseflow intensity has increased over the 68-year period. This is a classic indicator of climate change impacts on groundwater and soil saturation.

```{r setup, include=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(viridis)

# Load the processed yearly data
yearly_max <- rast("yearly_max_1945_2012.tif")
monthly_max <- rast("monthly_max_1945_2012.tif")
yearly_max_rcp45 <- rast("yearly_max_RCP45_1945_2099.tif")
monthly_max_rcp45<- rast("monthly_max_RCP45_1945_2099.tif")
yearly_max_rcp85 <- rast("yearly_max_RCP85_1945_2099.tif")
monthly_max_rcp85<- rast("monthly_max_RCP85_1945_2099.tif")

# 1. Calculate the regional average of the yearly maximums
regional_annual_max <- global(yearly_max, "mean", na.rm = TRUE)


nc_path <- "Gridded 1945-2012 Baseflow.nc"
bf_stack <- rast(nc_path)["BASEFLOW"]
# 2. Create a dataframe for plotting
df_trend <- data.frame(
  Year = 1945:2012,
  Max_Baseflow = regional_annual_max$mean
)

```

```{r 1.5}
# 3. Plot with a trend line (Loess smoothing)
ggplot(df_trend, aes(x = Year, y = Max_Baseflow)) +
  geom_line(color = "steelblue", alpha = 0.5) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "loess", color = "darkred", fill = "pink") +
  theme_minimal() +
  labs(title = "Average Annual Maximum Baseflow (1945-2012)",
       subtitle = "Aggregated across the Vancouver/PNW Region",
       y = "Baseflow (mm/day)",
       x = "Year")

#show the plot

```

---

## 2.Spatial Visualization (Snapshots)

Let’s visualize the hydrology data for specific dates to ensure the spatial patterns look correct.

```{r 2}
# Define dates to plot
target_dates <- as.Date(c("1945-01-01", "1972-05-01", "1972-07-01", "1997-06-01"))
actual_times <- time(bf_stack)

# Find indices matching these dates
indices <- which(as.Date(actual_times) %in% target_dates)

if (length(indices) > 0) {
  par(mfrow = c(2, 2)) # 2x2 Grid
  for (i in indices) {
    plot(bf_stack[[i]], 
         main = paste("Baseflow:", actual_times[i]),
         col = topo.colors(50),
         axes = FALSE) 
  }
  par(mfrow = c(1, 1)) # Reset
} else {
  print("Selected dates not found in the dataset.")
}


```

---

## 3. Interactive Maps for Key Dates

```{r 3}
library(terra)
library(leaflet)
n_wettest <- 2  # Number of wettest days to show
target_date <- "1945-07-15"  # Specific date to plot

# ============================================
# Helper Function
# ============================================
create_baseflow_map <- function(layer, title_text) {
  all_values <- terra::values(layer)
  actual_range <- range(all_values, na.rm = TRUE)
  domain_vals <- c(0, actual_range[2])
  
  pal <- leaflet::colorNumeric(c("#FFFFCC", "#41B6C4", "#0C2C84"), 
                                domain = domain_vals,
                                na.color = "transparent")
  
  leaflet::leaflet() %>%
    leaflet::addProviderTiles(leaflet::providers$CartoDB.Positron) %>%
    leaflet::addRasterImage(layer, 
                            colors = pal, 
                            opacity = 0.8,
                            project = FALSE,
                            maxBytes = 20 * 1024 * 1024) %>%
    leaflet::addLegend(pal = pal, 
                       values = all_values,
                       title = title_text)
}

# ============================================
# 1. Top N Wettest Days
# ============================================
# daily_means <- global(bf_stack, "mean", na.rm = TRUE)
# top_indices <- order(daily_means$mean, decreasing = TRUE)[1:n_wettest]
# top_dates <- time(bf_stack)[top_indices]
# 
# top_maps <- lapply(1:n_wettest, function(i) {
#   layer <- bf_stack[[top_indices[i]]]
#   map_title <- sprintf("Rank %d - %s", i, top_dates[i])
#   create_baseflow_map(layer, map_title)
# })
# 
# # Display all top maps
# htmltools::tagList(top_maps)

# ============================================
# 2. Specific Date Map
# ============================================
specific_date <- as.Date(target_date)
date_index <- which(time(bf_stack) == specific_date)

if(length(date_index) > 0) {
  cat("\n=== Specific Date Map ===\n")
  cat(sprintf("Date: %s (Day %d)\n", specific_date, date_index))
  #cat(sprintf("Mean flow: %.2f mm\n", daily_means$mean[date_index]))
  
  specific_layer <- bf_stack[[date_index]]
  create_baseflow_map(specific_layer, paste("Baseflow on", specific_date))
} else {
  cat("\n⚠ Date", target_date, "not found in dataset\n")
  cat("Available range:", as.character(min(time(bf_stack))), 
      "to", as.character(max(time(bf_stack))), "\n")
}
```




## 4.Point Extraction Daily

We extract a time series for a specific coordinate. To ensure we don’t pick an ocean pixel (NA), we programmatically find a valid land pixel.

```{r 4}

# first_layer <- bf_stack[[1]]
# valid_points <- as.points(first_layer, values=TRUE, na.rm=TRUE)
# 
# # 2. Pick a safe sample point (e.g., the 1000th valid point)
# # Ensure we don't go out of bounds if the map is small
# idx <- min(1000, length(valid_points))
# sample_point <- valid_points[idx]
# coords <- crds(sample_point)
# 
# print(paste("Extracting data for Lat:", round(coords[1,2], 2), 
#             "Lon:", round(coords[1,1], 2)))

# # 1. Find valid land coordinates (not NA) from the first layer
#using the first layer of the yearly max to find valid land coordinates
first_layer <- yearly_max[[1]]
valid_points <- as.points(first_layer, values=TRUE, na.rm=TRUE)

# 2. Pick a safe sample point (e.g., the 1000th valid point)
# Ensure we don't go out of bounds if the map is small
idx <- min(1000, length(valid_points))
sample_point <- valid_points[idx]
coords <- crds(sample_point)

print(paste("Extracting data for Lat:", round(coords[1,2], 2), 
            "Lon:", round(coords[1,1], 2)))
# 3. Extract time series for this location
point_vals <- terra::extract(bf_stack, sample_point, ID=FALSE) # returns a vector

# 4. Create dataframe
point_ts <- data.frame(
  Date = time(bf_stack),
  Baseflow = as.numeric(point_vals)
)

# 5. Plot
ggplot(point_ts, aes(x = Date, y = Baseflow)) +
  geom_line(color = "darkgreen", linewidth = 0.8) +
  theme_light() +
  labs(title = paste("Local Baseflow Series"), 
       subtitle = paste("Coordinates:", round(coords[1,1], 2), ",", round(coords[1,2], 2)),
       y = "Baseflow (mm)")

#make the data start from 1945, so filter the data frame and plot again
point_ts <- point_ts %>% filter(Date >= as.Date("1945-02-02") & Date <= as.Date("2012-12-31"))
ggplot(point_ts, aes(x = Date, y = Baseflow)) +
  geom_line(color = "darkgreen", linewidth = 0.8) +
  theme_light() +
  labs(title = paste("Local Baseflow Series (1945-2012)"), 
       subtitle = paste("Coordinates:", round(coords[1,1], 2), ",", round(coords[1,2], 2)),
       y = "Baseflow (mm)")

#make the plot wider and there are a lot x value (days)
library(plotly)
p <- ggplot(point_ts, aes(x = Date, y = Baseflow)) +
  geom_line(color = "darkgreen") +
  theme_light()

ggplotly(p) # This makes the plot interactive in your RStudio Viewer
```

---

## 5. Identifying the "Record Breaker" Year

Which year in the 1945–2012 record actually held the highest baseflow pixel-wise? This identifies specific historical storm events.

```{r 5}
# Find the index (year) of the maximum value for every pixel
year_of_max <- which.max(yearly_max)

# Convert indices to actual years
# If index 1 is 1945, then Year = Index + 1944
historic_max_years <- year_of_max + 1944

plot(historic_max_years, 
     main = "Year of Historic Maximum Baseflow",
     col = terrain.colors(68),
     axes = FALSE)

```

```{r 5.5}
library(terra)
library(leaflet)

# ============================================================
# 1. Prepare the Data (Same as before)
# ============================================================
# Ensure historic_max_years exists from your previous code
# Bin years into numeric decades (e.g., 1956 -> 1950)
decades_rast <- floor(historic_max_years / 10) * 10

# Get the range of decades for the domain boundaries
decade_range <- range(values(decades_rast), na.rm = TRUE)

# ============================================================
# 2. Create the Gradient Color Palette (Red -> Blue)
# ============================================================
# We switch to 'colorNumeric' for a continuous gradient.
# We define the exact colors: Red for low values, Blue for high values.
pal_gradient <- colorNumeric(
  palette = c("#D73027", "#4575B4"), # Using slightly nicer hex codes for Red and Blue
  # Alternatively, just use standard colors: palette = c("red", "blue"),
  domain = decade_range,
  na.color = "transparent"
)

# ============================================================
# 3. Create the Interactive Map
# ============================================================
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addRasterImage(
    decades_rast,
    colors = pal_gradient, # Use the new gradient palette
    opacity = 0.8,
    maxBytes = 20 * 1024 * 1024,
    group = "Trend Gradient"
  ) %>%
  
  addLegend(
    pal = pal_gradient,
    values = values(decades_rast), # Supply values so it knows the range
    title = "Record-Breaking Era",
    position = "bottomright",
    # Keep the 's' suffix for readability
    labFormat = labelFormat(suffix = "s"),
    opacity = 1
  )
```
---

## 6. Statistical Distribution (Ridge Plot)

Since you have the `monthly_max_1945_2012.tif`, you can visualize how the distribution of "maximums" shifts through the seasons.

---

## Summary of Findings to look for:

1. **Topography Influence:** You should notice the highest values () concentrated along the Coast Mountains and the Cascades, where orographic lift intensifies rainfall.
2. **The 1990s/2000s Shift:** Check Section 1's trend line; many PNW models show a slight intensification of extreme events in the latter half of the 20th century.
3. **Pixel Size:** Remember that your  pixels are large enough to encompass entire small sub-watersheds, meaning these baseflow values represent a significant volume of water moving into the stream network.

**Would you like me to help you write a script to extract the data for a specific city (like Vancouver or Seattle) and perform a Gumbel Distribution fit to calculate the 50-year and 100-year return levels?**