---
title: "Hydrologic Hotspot Analysis: Multi-Threshold Comparison"
author: "Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(terra)
library(viridis)
library(RColorBrewer)

```

## 1. Data Loading

We load the pre-calculated threshold maps for History and RCP 4.5. The time-series analysis (counting exceedances) has been performed externally.

```{r load-threshold-maps}
# Historical Thresholds
t90_hist <- rast("threshold_90_history.tif")
t95_hist <- rast("threshold_95_history.tif")
t99_hist <- rast("threshold_99_history.tif")

# RCP 4.5 Thresholds
t90_rcp45 <- rast("threshold_90_rcp45.tif")
t95_rcp45 <- rast("threshold_95_rcp45.tif")
t99_rcp45 <- rast("threshold_99_rcp45.tif")

# RCP 8.5 Thresholds (Future Work)
# t90_rcp85 <- rast("threshold_90_rcp85.tif")
# t95_rcp85 <- rast("threshold_95_rcp85.tif")
# t99_rcp85 <- rast("threshold_99_rcp85.tif")

head(t95_rcp45
     
     ,n=50)
```


## 2 Spatial Hotspot Analysis
We count how many times each cell exceeded its 95th percentile over the 140-year record.

```{r}

```

## 2. Spatial Visualization: History vs. RCP 4.5

Below we compare the spatial distribution of the 95th percentile (as the primary POT baseline) between the two periods.

```{r plot-95-comparison, fig.height=6, fig.width=10}
par(mfrow = c(1, 2))

plot(t95_hist, 
     main = "History: 95th Percentile Threshold", 
     col = viridis(100), 
     plg = list(title = "Value"))

plot(t95_rcp45, 
     main = "RCP 4.5: 95th Percentile Threshold", 
     col = viridis(100), 
     plg = list(title = "Value"))

```

## 3. Threshold Magnitude Shift

This reveals how much the "extreme" definition itself changes between the historical period and the RCP 4.5 projection.

```{r calculate-diffs}
# Magnitude Differences
diff_90 <- t90_rcp45 - t90_hist
diff_95 <- t95_rcp45 - t95_hist
diff_99 <- t99_rcp45 - t99_hist

# RCP 8.5 Magnitude Differences (Future Work)
# diff_90_85 <- t90_rcp85 - t90_hist
# diff_95_85 <- t95_rcp85 - t95_hist
# diff_99_85 <- t99_rcp85 - t99_hist

```

```{r plot-magnitude-shifts, fig.height=5, fig.width=15}
par(mfrow = c(1, 3))

plot(diff_90, main = "Δ 90th Pctl (RCP4.5 - Hist)", col = brewer.pal(11, "RdBu"))
plot(diff_95, main = "Δ 95th Pctl (RCP4.5 - Hist)", col = brewer.pal(11, "RdBu"))
plot(diff_99, main = "Δ 99th Pctl (RCP4.5 - Hist)", col = brewer.pal(11, "RdBu"))

```

---

### Analysis Notes

* **Sensitivity:** Comparing the 90, 95, and 99 thresholds allows you to see if "moderate extremes" (90th) are shifting differently than "rare extremes" (99th).
* **Hotspot Identification:** Regions where the difference maps show high positive values (dark red) indicate that the baseline for what constitutes an "extreme" event is rising significantly under climate change.
* **Alignment:** Ensure all `.tif` files share the same Coordinate Reference System (CRS). You can check this using `crs(t95_hist)`.
