
---
title: "Baseflow Data Processing Report (1945-2012)"
subtitle: "Temporal Aggregation and Spatial Resolution Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(terra)

```


## Overview

This report documents the analysis of gridded baseflow data from 1945 to 2012 for the Vancouver/Pacific Northwest region. The workflow calculates monthly and yearly maximum baseflow values from daily NetCDF data using an optimized incremental processing approach.

**Data Characteristics:**

- Variable: BASEFLOW (mm/day)
- Temporal Coverage: 1945-2012 (68 years)
- Spatial Resolution: 0.0625° × 0.0625° (1/16 degree grid)
- Coordinate System: WGS 84 (EPSG:4326)
- Region: Vancouver/Pacific Northwest

**Output Products:**

- Monthly maximum baseflow: 816 layers (68 years × 12 months)
- Yearly maximum baseflow: 68 layers (one per year)

## Spatial Resolution Details

The data is recorded on a 1/16 degree grid, where each pixel is exactly 0.0625° in both latitude and longitude.

**Physical dimensions near Vancouver (49.28°N):**

- **Latitude (Height)**: 0.0625° × 111 km ≈ **6.94 km**
  - This dimension remains constant regardless of latitude
  
- **Longitude (Width)**: 0.0625° × 111 km × cos(49.28°) ≈ **4.53 km**
  - This dimension decreases as you move northward due to meridian convergence
  
- **Pixel Area**: Approximately **31.4 km²** per pixel in the Vancouver region

- **Total Pixel Count**: The dataset contains 35,041 pixels per layer (calculated from the spatial extent and resolution).
## Processing Strategy

The analysis employs a two-stage aggregation strategy to optimize memory usage and computational efficiency:

**Stage A: Daily → Monthly Maximum**

This is the computationally intensive step, processing up to 31 daily layers per month.

**Stage B: Monthly → Yearly Maximum**

This is extremely fast because it only processes 12 monthly layers per year instead of 365 daily layers.

The data is processed in decade-long chunks (10-year groups) to manage memory constraints.

## save the final outputs as GeoTIFFs for easy sharing and visualization in GIS software.
The name of the output files are:
- `monthly_max_1945_2012.tif`
- `yearly_max_1945_2012.tif`

```{r data info}
# > print(final_yearly_max)
# class       : SpatRaster 
# size        : 367, 496, 68  (nrow, ncol, nlyr)
# resolution  : 0.0625, 0.0625  (x, y)
# extent      : -140, -109, 41.0625, 64  (xmin, xmax, ymin, ymax)
# coord. ref. : lon/lat WGS 84 (EPSG:4326) 
# source      : yearly_max_1945_2012.tif 
# names       :         1945,         1946,         1947,         1948,         1949,         1950, ... 
# min values  : 2.734829e-06, 2.734829e-06, 2.734829e-06, 2.734829e-06, 2.734829e-06, 2.734829e-06, ... 
# max values  : 1.680794e+02, 6.995869e+01, 7.105131e+01, 7.759936e+01, 8.262132e+01, 8.449879e+01, ... 
# > print(res(final_yearly_max))
# [1] 0.0625 0.0625
#The variable is BASEFLOW from a NetCDF file, representing the annual maximum 
#for each pixel from 1945–2012. The unit is mm/day. The region is Vancouver/PNW.


```
## 1. Introduction and Setup

This report documents the processing of gridded baseflow data from 1945 to 2012. The objective is to aggregate daily NetCDF data into **Monthly Maximum** and **Yearly Maximum** raster stacks.

To handle the large temporal range efficiently, we utilize `terra`'s memory management options and process the data in 10-year incremental chunks.

```{r load-data}
# 1. Setup and Memory Management
terraOptions(memmax = 0.4, progress = 10) 

# Define input path and load reference
nc_path <- "Gridded 1945-2012 Baseflow.nc"

# Note: We load the source lazily using rast()
bf_stack <- rast(nc_path)["BASEFLOW"]

# Get unique years to create processing groups
all_years <- unique(format(time(bf_stack), "%Y"))

# Split years into groups of 10 to optimize memory usage
year_groups <- split(all_years, ceiling(seq_along(all_years) / 10))

# Initialize storage lists
monthly_max_list <- list()
yearly_max_list <- list()

```

## 2. Optimized Incremental Processing

The following loop iterates through the defined year groups.

**Methodology:**

1. **Step A (Monthly Max):** We calculate the maximum daily value for every month. This is the most computationally intensive step.
2. **Step B (Yearly Max):** We derive the yearly maximum *from the monthly results*. This is significantly faster than reprocessing the daily data, as it reduces the operation from ~365 layers/year to just 12 layers/year.

```{r processing-loop, eval=FALSE}
# print("Starting Optimized Incremental Processing...")
# 
# for (i in seq_along(year_groups)) {
#   current_years <- year_groups[[i]]
#   print(paste0("Processing Group ", i, "/", length(year_groups), 
#                " (Years: ", min(current_years), "-", max(current_years), ")"))
#   
#   # 1. Subset Raw Data for the current time window
#   sub_stack <- bf_stack[[format(time(bf_stack), "%Y") %in% current_years]]
#   
#   # ---------------------------------------------------------
#   # STEP A: Calculate Monthly Max (The Heavy Lift)
#   # ---------------------------------------------------------
#   # Create index YYYY-MM
#   idx_monthly <- format(time(sub_stack), "%Y-%m")
#   
#   # Tapp 1: Raw Daily -> Monthly Max
#   chunk_monthly <- tapp(sub_stack, index = idx_monthly, fun = max, na.rm = TRUE)
#   monthly_max_list[[i]] <- chunk_monthly
#   
#   # ---------------------------------------------------------
#   # STEP B: Calculate Yearly Max (The Shortcut)
#   # ---------------------------------------------------------
#   # Extract year ("1945") from the monthly indices
#   unique_months <- unique(idx_monthly)
#   idx_yearly_from_month <- substr(unique_months, 1, 4)
#   
#   # Tapp 2: Monthly Max -> Yearly Max
#   chunk_yearly <- tapp(chunk_monthly, index = idx_yearly_from_month, fun = max, na.rm = TRUE)
#   yearly_max_list[[i]] <- chunk_yearly
#   
#   # ---------------------------------------------------------
#   # Cleanup
#   rm(sub_stack, chunk_monthly, chunk_yearly)
#   gc() 
# }
# 
# print("Loop finished. Ready to combine.")

```

## 3. Combining and Saving Results

The processed chunks are combined into final raster stacks and exported as GeoTIFFs.

### 3.1 Process Monthly Maximums

```{r save-monthly, eval=FALSE}
#final_monthly_max <- rast(monthly_max_list)
# writeRaster(final_monthly_max, "monthly_max_1945_2012.tif", overwrite = TRUE)
# print("Saved: monthly_max_1945_2012.tif")

```

### 3.2 Process Yearly Maximums

```{r save-yearly, eval=FALSE}
# final_yearly_max <- rast(yearly_max_list)
# 
# # Verify layer count matches year count
# if (nlyr(final_yearly_max) == length(all_years)) {
#   names(final_yearly_max) <- all_years
# } else {
#   warning("Layer count mismatch on yearly stack.")
# }
# 
# writeRaster(final_yearly_max, "yearly_max_1945_2012.tif", overwrite = TRUE)
# print("Saved: yearly_max_1945_2012.tif")

```

## 4. Verification and Spatial Analysis

### 4.1 Data Structure Verification

We perform a final check on the output dimensions to ensure the years 1945–2012 are fully represented.

```{r verification, eval=FALSE}
#read the tif files we just created
final_monthly_max <- rast("monthly_max_1945_2012.tif")
final_yearly_max <- rast("yearly_max_1945_2012.tif")
# Check structure
print(final_yearly_max)

# Print the last 10 years to verify temporal coverage
print(tail(names(final_yearly_max), 10))

```

### 4.2 Resolution and Grid Analysis

The dataset uses a WGS 84 Coordinate Reference System. The resolution is confirmed to be **1/16th degree**.

```{r spatial-check}
# Resolution check
res_val <- res(bf_stack)
print(paste("Resolution:", res_val[1], "degrees"))

# Coordinate System
print(crs(bf_stack, describe = TRUE)$name)

```

### 4.3 Physical Area Calculation (Vancouver Context)

Given the resolution of  (), we calculate the physical dimensions of a pixel for the region of interest (Vancouver/PNW, approx. latitude  N).

**1. Height (Latitude):**
The distance between degrees of latitude is approximately constant.

**2. Width (Longitude):**
The distance between degrees of longitude varies by the cosine of the latitude.

**3. Total Pixel Area:**
Therefore, each pixel in the Vancouver region covers an approximate area of:

### 4.4 Unit Verification

The variable `BASEFLOW` represents the annual maximum for each pixel.

```{r units}
# Check units of the raw stack
#show only first several layers to avoid overwhelming output

#print(paste("Data Units:", units(bf_stack)))
print(paste("Units of first layer:", units(bf_stack[[1]])))

```

## 5. Missing Value Analysis (QA/QC)
This section analyzes the integrity of the raw data by identifying NA (missing) values across three dimensions: total count, temporal distribution (by month/year), and spatial distribution (by pixel location).
```{r}
# library(terra)
# library(dplyr)
# library(knitr)
# 
# # 1. Define the Master Study Area Footprint
# # ---------------------------------------------------------
# # A pixel is "Inside" if it has > 0 valid observations over the entire 68 years.
# # This creates a single-layer mask where 1 = Study Area, NA = True Background.
# master_footprint <- sum(!is.na(final_monthly_max)) > 0
# master_footprint[master_footprint == 0] <- NA 
# 
# # 2. Find Gaps (Holes)
# # ---------------------------------------------------------
# # We calculate how many layers are missing for each pixel THAT IS INSIDE the footprint.
# # is.na(final_monthly_max) returns 1 for NA, 0 for value.
# # We multiply by the master_footprint to ignore the "Ocean/Background".
# gap_counts <- sum(is.na(final_monthly_max)) * master_footprint
# 
# # 3. Spatial Analysis: Locating the "Specks"
# # ---------------------------------------------------------
# holes_df <- as.data.frame(gap_counts, xy = TRUE)
# colnames(holes_df)[3] <- "missing_months"
# 
# # Real holes are pixels inside the footprint that have at least 1 missing month
# internal_specks <- holes_df %>%
#   filter(!is.na(missing_months)) %>%
#   filter(missing_months > 0) %>%
#   arrange(desc(missing_months))
# 
# # 4. Visualization
# # ---------------------------------------------------------
# # Plot the footprint in light grey and the holes in red
# plot(master_footprint, col="grey90", main="Internal Data Gaps (Red Specks)", legend=FALSE)
# if(nrow(internal_specks) > 0) {
#   # Convert speck coordinates to points for clear visibility on the map
#   points(internal_specks$x, internal_specks$y, col="red", pch=20, cex=0.5)
# }
# 
# # 5. Output Results
# # ---------------------------------------------------------
# cat("### QA Summary: Internal Gaps\n")
# if (nrow(internal_specks) == 0) {
#   cat("Success: No internal data gaps (specks) found within the study area.\n")
# } else {
#   cat("Detected", nrow(internal_specks), "pixels with intermittent missing data.\n")
#   kable(head(internal_specks, 10), caption = "Top 10 pixels with the most frequent internal gaps.")
#}
```