---
title: "Non-stationary conditional extremes for BC streamflow: the code report (v5, consolidated)"
author: "Alex Lin (supervisor: Natalia Nolde)"
date: "August 18, 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
params:
  start_year: 1969     # meeting 62: all of 1968 is discarded before anything runs
  spacing_yr: 5        # trend-spline knot spacing (v8 Sec 4.2, p. 7)
  doy_k: 12            # cyclic day-of-year basis size (threshold); 8 at the GP stage
  tau_star: 0.95       # working threshold level (v8 Sec 2.1, p. 3)
  kappa: 0.95          # dependence threshold level on the Laplace scale (v8 Sec 3, p. 3)
  cond_level: 0.99     # conditioning level for the three functionals (v8 Sec 6, eq. 14, p. 16)
  band_lo: 0.95        # P_band lower bound: the day's marginal quantile level
  band_hi: 0.995       # P_band upper bound
  B_boot: 25           # knit-scale bootstrap; paper scale is 2000 with selections inside
  n_mc: 10000          # Monte Carlo generation size (v8 Sec 6, p. 16)
  refit_ladders: false # false = reuse cache_ladder_*.rds; true = force refit
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9.5, fig.height = 5.5, dpi = 110)
set.seed(20260818)
library(quantreg)
library(splines)

## software routes: packages are production where installed; the project
## fitters are the fallback and, where both run, the consistency check.
## qgam is the threshold production route (decision of Aug 18, on the
## qgam papers); evgam the GP route; texmex/mgcv the dependence checks.
has <- c(qgam     = requireNamespace("qgam",     quietly = TRUE),
         evgam    = requireNamespace("evgam",    quietly = TRUE),
         extRemes = requireNamespace("extRemes", quietly = TRUE),
         texmex   = requireNamespace("texmex",   quietly = TRUE),
         mgcv     = requireNamespace("mgcv",     quietly = TRUE))
if (has["qgam"]) library(qgam)   # qgam's formulas resolve on the search path
```

This report is the working pipeline for the paper *Non-stationary modelling
of stream flow extremes under climate change scenarios*: one pair first ---
Case 4, BRIDG given NTHMB, with Case 1 (BARRM given BARRS) as the
asymptotic-dependence anchor (v8 \S1, p. 2; \S5, p. 12). Every reference of
the form "v8 \S$x$, eq. ($n$), p. $y$" points into *my current understanding
of the model*, version 8. Four design facts orient everything below.

**The working scale is Laplace** (v8 \S3, eq. 6, p. 4; Keef et al. 2013, eq.
2.1): both tails exactly exponential, so one parametric form covers positive
and negative association with $a \in [-1,1]$ (v8 \S5, eq. 8, p. 11), and the
Keef parameter constraints police extrapolation (v8 \S5, eq. 9, p. 13).

**The margins carry the seasonal cycle alongside the trend.** The marginal
diagnostic of v8 \S3 (p. 4) failed in review --- the PIT showed a clear
seasonal shape --- so the repair v8 \S4.2 (p. 7) names is adopted: the
single-$t$ basis is swapped for one that contains a shared periodic
day-of-year cycle next to the trend spline, coefficients, penalty and
fitting unchanged. The threshold now moves with the seasons within each
year, and "roughly 5\% above the threshold" becomes checkable *locally* ---
per season and per decade --- not just pooled. One consequence is flagged
for discussion: with a seasonal threshold, "extreme" means extreme relative
to the season, so a winter exceedance is a modest absolute flow; the annual
return level (v8 \S6, eq. 12, p. 14) is unaffected, but the reading of the
dependence-stage conditioning event changes and should be agreed explicitly.

**Data are cleaned before anything is fitted.** All of 1968 is discarded;
the forensics section hunts the repeated values, the low-flow floor and the
winter quality problems, and reports evidence --- which days, which values,
which months --- so the exclusion decisions can be made deliberately and
documented in the paper. No exclusion beyond 1968 is applied silently.

**Packages are production where they exist** (v8 \S4.5, p. 9): qgam for the
threshold ladder (automatic smoothing; its three recorded costs --- no
non-crossing constraint, iid calibration theory, care at extreme rungs ---
are v8 \S4.4, p. 9, and each is handled below), evgam for the GP tail,
extRemes for the extremal index, texmex and mgcv as the dependence checks.
The dependence fitter itself is project code: no package fits the
non-stationary Heffernan--Tawn model. The routes table states the intent;
each stage table below reports the route actually used after any fallback.

```{r routes}
route <- data.frame(
  stage = c("threshold ladder", "GP tail", "extremal index",
            "dependence fitter", "dependence checks"),
  production = c(
    ifelse(has["qgam"], "qgam/mqgam (seasonal + trend)", "project pen_rq (fallback)"),
    ifelse(has["evgam"], "evgam (gpd, seasonal + trend)", "project fit_gp (fallback)"),
    ifelse(has["extRemes"], "extRemes::extremalindex", "project ferro_segers (fallback)"),
    "project ht fitter with Keef constraints (no package exists)",
    paste0("tier 1 texmex: ", ifelse(has["texmex"], "on", "off"),
           " | tier 2 mgcv: ", ifelse(has["mgcv"], "on", "off"))))
knitr::kable(route, caption = "Software routes this knit -- the INTENT. The
stage tables below each report the route actually used after any tryCatch
fallback; trust those, not this one.")
```

# Data, axis, integrity gates

The file is the definitive RCP 4.5 export; the axis is its extent *after*
the 1968 drop (v8 \S1, p. 1: RCP 4.5 alone; meeting 62: discard 1968).

```{r data}
raw <- read.csv("daily_sf_bf_rcp45_only_1945_2099_9avg.csv",
                colClasses = c("character", "character", "numeric", "numeric"))
names(raw) <- c("station", "date", "sf", "bf")
raw$date <- as.Date(raw$date)
raw$year <- as.integer(format(raw$date, "%Y"))

## integrity gates on the RAW file, before the 1968 drop
dup_rows <- sum(duplicated(raw[, c("station", "date")]))
cat("duplicated (station, date) rows in the file:", dup_rows, "\n")
full_seq <- seq(min(raw$date), max(raw$date), by = "day")
miss <- sapply(split(raw$date, raw$station),
               function(d) length(setdiff(full_seq, d)))
cat("missing dates per station (0 everywhere = continuous):",
    paste(range(miss), collapse = "-"), "\n")
cat("NA counts: sf", sum(is.na(raw$sf)), "| bf", sum(is.na(raw$bf)), "\n")
cat("bf <= 0 days, worst stations:\n")
bfneg <- sort(tapply(raw$bf <= 0, raw$station, sum), decreasing = TRUE)
print(head(bfneg[bfneg > 0], 5))

keep1968 <- raw[raw$year < params$start_year, ]   # kept aside for forensics
d68 <- keep1968
raw <- raw[raw$year >= params$start_year, ]

a0 <- min(raw$date); a1 <- max(raw$date)
dates <- sort(unique(raw$date)); N <- length(dates)
years <- as.integer(format(dates, "%Y"))
months <- as.integer(format(dates, "%m"))
doy <- as.integer(format(dates, "%j"))
tt <- as.numeric(dates - a0) / as.numeric(a1 - a0)
span_yr <- as.numeric(a1 - a0) / 365.25

get_series <- function(st, var, df = raw) {
  d <- df[df$station == st, ]; d[[var]][order(d$date)]
}
C_st <- "NTHMB"; X_st <- "BRIDG"          # Case 4: X | C (v8 Sec 1, p. 2)
A_C <- "BARRS"; A_X <- "BARRM"            # Case 1 anchor (v8 Sec 5, p. 12)
xC <- get_series(C_st, "sf"); xX <- get_series(X_st, "sf")
xB <- get_series(X_st, "bf")              # Structure-B conditioning variable

stopifnot(length(xC) == N, length(xX) == N, !anyNA(c(xC, xX, xB)),
          all(xC > 0), all(xX > 0))
cat("Axis after cleaning gate:", format(a0), "to", format(a1), "-", N,
    "days,", round(span_yr, 1), "years\n")
```

# Data forensics (meeting 62, the first problem)

Three artifacts motivated this section, seen on the previous knit's figures:
a flat run in the pooled QQ plots ending in a jump near sample value $-2.4$,
and two straight line segments hugging the axes of the joint pseudosample
panel. Three candidate causes, in decreasing order of code-vs-data:

(i) *a transform artifact*: the old transform clamped every day whose
fitted probability fell below $10^{-5}$ to that single number, and
$-\log(-\log 10^{-5}) = -2.44$ on the Gumbel scale --- exactly the observed
floor. All days far enough below the bottom ladder rung collapsed onto one
point, which draws a flat QQ run ending in a jump, and straight lines in
any joint scatter. Section 5's rebuilt transform removes the clamp; its
instrumentation counts the affected days so the artifact is quantified, not
just repaired.

(ii) *repeated identical values in the data*: ties --- from winter freezing
floors, detection limits, or values entered more than once --- put genuine
point masses in the sample. Ties survive any honest transform (they are
data, not code), and this section finds them: which values, which months,
which years.

(iii) *the trend-only margins*: a threshold that cannot bend within a year
maps winter days systematically low, so the bottom of the pooled sample is
mostly December--February and any low-flow pathology concentrates there.
Section 3's seasonal threshold removes this confound; whatever remains
after it is data, not design.

## Repeated exact values: which, how often, when

```{r dupvalues}
series_list <- list(`NTHMB sf` = xC, `BRIDG sf` = xX, `BRIDG bf` = xB)
month_lab <- function(m) paste(sort(unique(month.abb[m])), collapse = " ")
top_ties <- function(x, k = 10) {
  tb <- head(sort(table(x), decreasing = TRUE), k)
  do.call(rbind, lapply(names(tb), function(v) {
    i <- which(x == as.numeric(v))
    data.frame(value = as.numeric(v), n_days = length(i),
               years = paste(range(years[i]), collapse = "-"),
               n_years = length(unique(years[i])),
               months = month_lab(months[i]))
  }))
}
for (nm in names(series_list)) {
  cat("\n----", nm, ": the ten most repeated exact values ----\n")
  print(top_ties(series_list[[nm]]), row.names = FALSE)
  x <- series_list[[nm]]
  cat(sprintf("days whose value occurs >= 10 times: %d of %d (%.1f%%)\n",
      sum(x %in% as.numeric(names(which(table(x) >= 10)))), N,
      100 * mean(x %in% as.numeric(names(which(table(x) >= 10))))))
}
```

**Reading the table.** One block per series. `value` is an exact repeated
number; `n_days` how many days carry it; `years`/`n_years` when; `months`
where in the year. What to look for: a value repeated hundreds of times in
DJF across many years is a winter measurement floor (freezing); a value
repeated only within one or two years is closer to a data-entry duplicate
--- the professor's question "which values are these, and is it only that
year" is answered by these columns. Large `n_days` at the series minimum is
the floor that drew the QQ jump.

## Distinct values per year, and runs of identical days

```{r distinctruns, fig.height=4.2}
op <- par(mfrow = c(1, 3), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
yrs_u <- sort(unique(years))
for (nm in names(series_list)) {
  x <- series_list[[nm]]
  nd <- sapply(yrs_u, function(y) length(unique(x[years == y])) /
                                   sum(years == y))
  plot(yrs_u, nd, type = "h", lwd = 2, col = "steelblue4", ylim = c(0, 1),
       xlab = "year", ylab = "share of distinct values", main = nm)
  abline(h = 1, lty = 3, col = "grey60")
}
par(op)

run_table <- function(x, minlen = 3) {
  r <- rle(x)
  i_end <- cumsum(r$lengths); i_start <- i_end - r$lengths + 1
  keep <- r$lengths >= minlen
  data.frame(start = dates[i_start[keep]], length = r$lengths[keep],
             value = r$values[keep], month = months[i_start[keep]])
}
for (nm in names(series_list)) {
  rt <- run_table(series_list[[nm]])
  cat("\n----", nm, sprintf(": %d runs of >= 3 identical consecutive days ----\n",
      nrow(rt)))
  if (nrow(rt) > 0) {
    print(head(rt[order(-rt$length), c("start", "length", "value")], 8),
          row.names = FALSE)
    print(table(month = rt$month))
  }
}
```

**Reading the figure and tables.** Each panel: x-axis the year, bar height
the share of that year's 365 days carrying distinct values --- 1.0 means no
two days of the year had exactly equal flow; visibly short bars mean the
year is full of repeats. The runs table shows consecutive identical days
--- a plateau: rivers vary daily, so a value frozen for a week is an
instrument or post-processing signature, not hydrology. The month row of
the table says where plateaus live; a DJF concentration is the freezing
hypothesis confirmed in this series.

## The floor: days at or near the minimum

```{r floor}
for (nm in names(series_list)) {
  x <- series_list[[nm]]
  at_min <- sum(x == min(x)); near <- sum(x <= min(x) * 1.01)
  cat(sprintf("\n%s: min = %.4f | days exactly at min: %d | within 1%%: %d\n",
      nm, min(x), at_min, near))
  if (at_min > 1) {
    i <- which(x == min(x))
    cat("  at-min days by month:\n"); print(table(month = months[i]))
    tab <- sort(table(years[i]), decreasing = TRUE)
    cat("  worst years:", paste(names(tab)[1:min(5, length(tab))],
        "(", tab[1:min(5, length(tab))], ")"), "\n")
  }
}
```

**Reading the output.** A minimum carried by one day is ordinary; a minimum
carried by dozens of days is a hard floor --- a detection limit or a frozen
gauge --- and those are exactly the days the old clamp mapped to $-2.44$.
Their month distribution says whether the floor is a winter phenomenon;
their year distribution says whether it is confined to particular years
(then a targeted exclusion is defensible) or endemic (then the handling is
a modelling decision to discuss).

## Winter windows, by eye

```{r winterwindows, fig.height=4.5}
rtX <- run_table(xX); rtC <- run_table(xC)
worst <- unique(c(format(rtX$start[order(-rtX$length)], "%Y"),
                  format(rtC$start[order(-rtC$length)], "%Y")))
worst <- as.integer(worst)
worst <- worst[!is.na(worst) & worst < max(years)]
worst <- unique(c(worst, yrs_u[2], yrs_u[5]))[1:2]
op <- par(mfrow = c(1, 2), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
for (w in worst) {
  k <- dates >= as.Date(sprintf("%d-11-01", w)) &
       dates <= as.Date(sprintf("%d-04-30", w + 1))
  matplot(dates[k], cbind(xC[k], xX[k]), type = "l", lty = 1, lwd = 1.6,
          col = c("steelblue4", "firebrick3"), xlab = "date",
          ylab = "daily flow", main = sprintf("winter %d-%d", w, w + 1))
  legend("topright", c("NTHMB sf", "BRIDG sf"), lwd = 2, bty = "n",
         col = c("steelblue4", "firebrick3"), cex = 0.8)
}
par(op)
```

**Reading the figure.** Each panel is one November--April window --- the two
winters holding the longest identical-value plateaus, picked automatically.
The x-axis is the date, the y-axis raw daily flow, one colour per station.
What a measurement problem looks like: a series that locks onto a perfectly
horizontal segment for days or weeks (a frozen or floored gauge), steps
between a few discrete levels, or drops to a repeated identical minimum ---
against the smooth recessions hydrology produces. These panels are the
"identifiable by eye" evidence the meeting asked for; more winters can be
inspected by changing `worst`.

## 1968 against its neighbours

```{r audit1968}
aud <- function(df, st, var) {
  x <- df[[var]][df$station == st]
  c(n = length(x), distinct = length(unique(x)),
    share_repeated = round(mean(x %in% as.numeric(names(which(table(x) >= 2)))), 3))
}
cmp <- rbind(`1968` = aud(d68, X_st, "sf"),
             do.call(rbind, lapply(params$start_year:(params$start_year + 4),
               function(y) aud(raw[raw$year == y, ], X_st, "sf"))))
rownames(cmp)[-1] <- params$start_year:(params$start_year + 4)
knitr::kable(cmp, caption = sprintf("BRIDG sf: 1968 (dropped) against the
first kept years -- sample size, distinct values, share of days whose value
repeats within the year. 1968 is excluded from the axis by decision
(meeting 62); this table records the evidence beside the decision."))
```

# Stage 1a: the seasonal moving threshold (v8 \S2.1, eq. 3, p. 3; repair of \S4.2, p. 7)

The threshold is the conditional quantile curve $u(t;\tau)$, now with two
components: a shared periodic day-of-year cycle (one shape, learned from
all years at once --- a cyclic basis closes December 31 onto January 1) and
the slow trend spline across the axis. Production route: `qgam`/`mqgam` ---
smoothing selected automatically by approximate marginal likelihood, the
loss calibrated so intervals hold roughly their coverage (v8 \S4.4, p. 9),
and the location-scale formula letting the loss bandwidth follow the
strongly seasonal variance of flow. The ladder is fitted per level with no
non-crossing constraint --- qgam's recorded first cost --- so non-crossing
is audited and repaired by monotone rearrangement (Chernozhukov,
Fern\'andez-Val \& Galichon 2010), with the crossing count printed: a
near-zero count is the empirical licence for the qgam ladder. Fallback +
consistency route: the project penalized-`rq` fitter on the same seasonal +
trend design, $\lambda_u$ by year-block cross-validation (v8 \S4.4, p. 9).

```{r qrmachinery}
## ladder levels: low rungs added so the transform's lower tail has anchors
taus <- c(0.01, 0.025, 0.05, seq(0.10, 0.90, by = 0.10), params$tau_star)
K <- length(taus)

make_basis <- function(tt, spacing_yr, span_yr) {
  n_int <- max(1, floor(span_yr / spacing_yr) - 1)
  interior <- seq_len(n_int) / (n_int + 1)
  bs(tt, knots = interior, degree = 3, intercept = TRUE,
     Boundary.knots = c(0, 1))
}
harmonics <- function(doy, n_harm = 3) {
  do.call(cbind, lapply(seq_len(n_harm), function(k)
    cbind(sin(2 * pi * k * doy / 366), cos(2 * pi * k * doy / 366))))
}
Btrend <- make_basis(tt, params$spacing_yr, span_yr)
H <- harmonics(doy)
Xseas <- cbind(Btrend, H)                 # seasonal + trend design, fallback
p_tr <- ncol(Btrend)
cat("design: trend", p_tr, "+ harmonics", ncol(H), "columns\n")

year_block_folds <- function(years, K = 5) {
  yrs <- sort(unique(years))
  fold_of_year <- setNames(rep(seq_len(K), length.out = length(yrs)), yrs)
  fold_of_year[as.character(years)]
}
folds <- year_block_folds(years, K = 5)

pinball <- function(r, tau) sum(pmax(tau * r, (tau - 1) * r))
Dtr <- cbind(diff(diag(p_tr)), matrix(0, p_tr - 1, ncol(H)))
pen_rq_seas <- function(y, X, tau, lambda) {
  Xa <- rbind(X, lambda * Dtr, -lambda * Dtr)
  ya <- c(y, rep(0, 2 * nrow(Dtr)))
  rq.fit.fnb(Xa, ya, tau = tau)$coefficients
}
lam_grid <- 10^(0:4)

fit_ladder_own <- function(x) {
  cvs <- sapply(lam_grid, function(lam) {
    mean(sapply(sort(unique(folds)), function(k) {
      tr <- folds != k
      g <- pen_rq_seas(x[tr], Xseas[tr, , drop = FALSE], params$tau_star, lam)
      pinball(x[!tr] - Xseas[!tr, , drop = FALSE] %*% g, params$tau_star) / sum(!tr)
    }))
  })
  lam_u <- lam_grid[which.min(cvs)]
  G <- sapply(taus, function(tau) pen_rq_seas(x, Xseas, tau, lam_u))
  U <- Xseas %*% G
  ncross <- sum(apply(U, 1, function(r) any(diff(r) <= 0)))
  msize <- 0
  if (ncross > 0) {
    Us <- t(apply(U, 1, sort))
    msize <- max(abs(Us - U)) / sd(x)   # size of the largest repair
    U <- Us
  }
  list(U = U, lam_u = lam_u, ncross = ncross, msize = msize, route = "own")
}

fit_ladder_qgam <- function(x) {
  tryCatch({
    d <- data.frame(y = x, tt = tt, doy = doy)
    fm <- list(y ~ s(doy, bs = "cc", k = params$doy_k) + s(tt, bs = "cr", k = 25),
               ~ s(doy, bs = "cc", k = 8))
    fitM <- mqgam(fm, data = d, qu = taus,
                  argGam = list(knots = list(doy = c(0.5, 366.5))))
    U <- sapply(taus, function(q) as.numeric(qdo(fitM, q, predict, newdata = d)))
    ncross <- sum(apply(U, 1, function(r) any(diff(r) <= 0)))
    msize <- 0
    if (ncross > 0) {
      Us <- t(apply(U, 1, sort))
      msize <- max(abs(Us - U)) / sd(x)
      U <- Us
    }
    list(U = U, lam_u = NA, ncross = ncross, msize = msize, route = "qgam")
  }, error = function(e) {
    cat("NOTE: qgam branch failed (", conditionMessage(e),
        ") - project fallback used\n", sep = "")
    fit_ladder_own(x)
  })
}
fit_ladder <- function(x) if (has["qgam"]) fit_ladder_qgam(x) else fit_ladder_own(x)

fit_ladder_cached <- function(x, tag) {
  f <- sprintf("cache_ladder_%s_%s_%d_%.0f.rds", tag,
               ifelse(has["qgam"], "qgam", "own"), N, sum(x))
  if (!params$refit_ladders && file.exists(f)) {
    out <- readRDS(f)
    cat("ladder", tag, "loaded from cache:", f, "\n")
    return(out)
  }
  out <- fit_ladder(x); saveRDS(out, f)
  cat("ladder", tag, "fitted and cached:", f, "\n")
  out
}
```

```{r thresholdfit}
mC <- fit_ladder_cached(xC, "C"); mX <- fit_ladder_cached(xX, "X")
mB <- fit_ladder_cached(xB, "B")
knitr::kable(data.frame(
  series = c("NTHMB sf (C)", "BRIDG sf (X)", "BRIDG bf (B)"),
  route = c(mC$route, mX$route, mB$route),
  crossing_days_repaired = c(mC$ncross, mX$ncross, mB$ncross),
  largest_repair_sd = round(c(mC$msize, mX$msize, mB$msize), 4)),
  caption = "Threshold ladders: route actually used, and the non-crossing
  audit. Rungs are fitted independently (qgam's recorded first cost), so
  crossings can occur -- typically hairline overlaps among the
  tightly-spaced low rungs, which the monotone rearrangement repairs
  (Chernozhukov et al. 2010: the repair never worsens the estimate). The
  count says how often; the last column says how BIG the largest repair
  was, in sd units of the series -- a large count with a tiny magnitude is
  cosmetic, a large magnitude localizes a real fitting problem.")

## consistency check where the package route ran (project-brief discipline)
if (mX$route == "qgam") {
  own <- fit_ladder_own(xX)
  gap <- mean(abs(own$U[, K] - mX$U[, K])) / sd(xX)
  cat(sprintf("qgam vs project top-rung mean gap: %.4f sd units (different
  smoothing selection and seasonal bases; reported, not asserted)\n", gap))
} else if (has["qgam"]) {
  cat("WARNING: qgam is installed but its branch did not produce the ladder
  (see NOTE above) - no qgam-vs-project gap to report\n")
}
```

## The threshold seen at year scale (meeting 62, the second problem)

```{r thresholdplot, fig.height=4.5}
op <- par(mfrow = c(1, 3), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
show <- c(which.min(abs(taus - 0.05)), which.min(abs(taus - 0.5)), K)
lcol <- c("grey40", "steelblue3", "firebrick3")
for (s in list(list(xC, mC, "NTHMB sf (C)"), list(xX, mX, "BRIDG sf (X)"),
               list(xB, mB, "BRIDG bf (B)"))) {
  plot(dates, s[[1]], pch = ".", col = "grey80", xlab = "year",
       ylab = "daily flow", main = s[[3]])
  for (j in seq_along(show))
    lines(dates, s[[2]]$U[, show[j]], lwd = 1.2, col = lcol[j])
  legend("topleft", paste0("tau = ", taus[show]), col = lcol, lwd = 2,
         bty = "n", cex = 0.8)
}
par(op)
```

**Reading the figure.** One panel per series; x-axis calendar time, y-axis
daily flow, grey dots the raw series. The curves are three rungs of the
ladder: $\tau = 0.05$ (lower envelope), $\tau = 0.5$ (the moving median ---
the closest analogue of a location curve $\mu(t)$ in this threshold-based
route), and $\tau^{*} = 0.95$, the working threshold: the GP tail is fitted
to overshoots above the red curve. Unlike a trend-only fit, these curves
now oscillate within each year --- the seasonal cycle --- while their
year-to-year drift carries the climate signal. At this compression the
seasonal band looks solid; the next figure zooms in.

```{r thresholdwindows, fig.width=9.5, fig.height=4}
xCbf <- get_series(C_st, "bf")
mCbf <- fit_ladder_cached(xCbf, "Cbf")   # NTHMB baseflow ladder (one-time fit)
ser4 <- list(list(xC, mC, "NTHMB sf"), list(xX, mX, "BRIDG sf"),
             list(xCbf, mCbf, "NTHMB bf"), list(xB, mB, "BRIDG bf"))
set.seed(62)                              # reproducible window draw
for (s in ser4) {
  for (zy in sort(sample(min(years):(max(years) - 1), 5))) {
    k <- years %in% c(zy, zy + 1)
    sh <- mean(s[[1]][k] > s[[2]]$U[k, K])
    plot(dates[k], s[[1]][k], pch = 16, cex = 0.4, col = "grey60",
         xlab = "date", ylab = paste(s[[3]], "daily flow"),
         main = sprintf("%s %d-%d | share above u(t; %.2f): %.3f",
                        s[[3]], zy, zy + 1, params$tau_star, sh))
    lines(dates[k], s[[2]]$U[k, K], lwd = 2, col = "firebrick3")
    lines(dates[k], s[[2]]$U[k, which.min(abs(taus - 0.5))], lwd = 2,
          col = "steelblue3")
    legend("topleft", c(sprintf("u(t; %.2f)", params$tau_star), "median rung"),
           lwd = 2, col = c("firebrick3", "steelblue3"), bty = "n", cex = 0.85)
  }
}
```

**Reading the figures.** Twenty separate panels: for each of the four series
(both stations, streamflow and baseflow), five randomly drawn two-year
windows spread over the axis. Dots are daily values; red is the working
threshold $u(t;\tau^{*})$, blue the moving median. The threshold should ride
the seasonal wave --- high through the freshet, low through winter --- and
the number in each title is that window's own share of days above red.
Target 0.05; a two-year window holds about 730 days, so roughly 0.03--0.07
is sampling noise, while a title far outside that range localizes a
miscalibrated stretch (compare with the season-by-era table). The windows
are drawn with a fixed seed, so every knit shows the same ten years; change
`set.seed(62)` to draw a different sample.

## Local exceedance proportions (the "5\% locally" check)

```{r localexceed, fig.height=4}
seas_of <- function(m) c("DJF","DJF","MAM","MAM","MAM","JJA",
                         "JJA","JJA","SON","SON","SON","DJF")[m]
seas <- seas_of(months)
era <- cut(years, breaks = seq(min(years), max(years) + 22, by = 22),
           right = FALSE, dig.lab = 4)
excX <- xX > mX$U[, K]
tab <- round(tapply(excX, list(seas, era), mean), 3)
knitr::kable(tab, caption = sprintf("BRIDG: share of days above u(t; %.2f)
by season and era. Target %.2f in every cell -- locally, not just pooled
(meeting 62). A cell far below target says the threshold missed that
season's level in that era.", params$tau_star, 1 - params$tau_star))

sh <- sapply(yrs_u, function(y) mean(excX[years == y]))
plot(yrs_u, sh, type = "h", lwd = 2, col = "steelblue4",
     xlab = "year", ylab = "share above threshold",
     main = sprintf("BRIDG: yearly share above u(t; %.2f)", params$tau_star))
abline(h = 1 - params$tau_star, col = "firebrick3", lwd = 2, lty = 2)
```

**Reading the figure and table.** Table: rows are seasons, columns eras;
each cell the share of that season-era's days above the working threshold.
The design target is `r 1 - params$tau_star` everywhere --- the moving
threshold's defining property (v8 \S3, p. 4), now demanded per season as
well as per era. Figure: the same share per calendar year against the
dashed target line; wandering above and below is sampling noise, a
sustained drift is a threshold failure. This is also qgam's own recommended
bias check for its smoothed loss --- one table serves both purposes. If the
cells miss the target in a season-by-era *pattern* --- one season too low
early and too high late --- the additive design (one seasonal shape shared
by every year, plus a trend) is too rigid, and the next design step is the
tensor interaction `te(doy, tt)`: the seasonal shape itself allowed to
drift, which the qgam route fits with one formula change.

## The tensor experiment: does the seasonal shape itself drift?

```{r tethreshold}
if (has["qgam"]) {
  f_te <- sprintf("cache_uX_te_qgam_%d_%.0f.rds", N, sum(xX))
  u_te <- tryCatch({
    if (!params$refit_ladders && file.exists(f_te)) readRDS(f_te) else {
      d <- data.frame(y = xX, tt = tt, doy = doy)
      fm <- list(y ~ te(doy, tt, bs = c("cc", "cr"), k = c(10, 15)),
                 ~ s(doy, bs = "cc", k = 8))
      kn <- list(knots = list(doy = c(0.5, 366.5)))
      fte <- qgam(fm, data = d, qu = params$tau_star, argGam = kn)
      u1 <- as.numeric(predict(fte, newdata = d))
      saveRDS(u1, f_te); u1
    }
  }, error = function(e) {
    cat("NOTE: te branch failed (", conditionMessage(e), ")\n", sep = ""); NULL })
  if (!is.null(u_te)) {
    tab_te <- round(tapply(xX > u_te, list(seas, era), mean), 3)
    knitr::kable(tab_te, caption = "BRIDG, tensor te(doy, tt) threshold at
    0.95: the season x era exceedance table again. If the additive design's
    first-era misses shrink toward 0.05 here, the seasonal shape itself
    drifts across the century and the tensor becomes the production design.")
  }
} else cat("qgam absent in this environment - te experiment skipped\n")
```

## Move one at the threshold (v8 \S4.6, p. 10)

```{r thrmove1}
thr_move1 <- function(x, B = 30) {
  Xlin <- cbind(1, tt, H)
  sl <- rq.fit.fnb(Xlin, x, tau = params$tau_star)$coefficients[2]
  yrs <- unique(years)
  reps <- replicate(B, {
    keep <- years %in% sample(yrs, replace = TRUE)
    rq.fit.fnb(Xlin[keep, ], x[keep], tau = params$tau_star)$coefficients[2]
  })
  c(slope_tt = sl, se = sd(reps), z = sl / sd(reps))
}
knitr::kable(round(rbind(`NTHMB sf` = thr_move1(xC), `BRIDG sf` = thr_move1(xX),
                         `BRIDG bf` = thr_move1(xB)), 3),
             caption = "Move one at the threshold: the trend slope of the
             linear-in-t rung (seasonal harmonics included in both models,
             so the slope is the trend, not the seasons) against its
             year-block bootstrap SE. A constant winning is a legitimate
             outcome (v8 Sec 4.6, p. 10).")
```

# Stage 1b: GP tail (v8 \S2.2, eq. 4, p. 3)

The scale $\sigma(t)$ gets the same seasonal + trend structure as the
threshold --- with a seasonal threshold, winter exceedances exist and their
overshoots are far smaller than freshet overshoots, so a seasonal scale is
not optional. The shape $\xi$ stays constant per series (decided; v8 \S2.2,
p. 3) and is printed --- the number to quote when asked. Production route:
evgam (`family = "gpd"`, REML smoothing); fallback + check: the project
penalized likelihood on the seasonal + trend design.

```{r gpmachinery}
gp_nll <- function(par, y, Xg, lambda = 0, Dg = NULL) {
  pg <- ncol(Xg)
  gam <- par[1:pg]; xi <- par[pg + 1]
  sig <- exp(as.vector(Xg %*% gam))
  z <- 1 + xi * y / sig
  if (any(z <= 0)) return(1e10)
  nll <- sum(log(sig) + (1 + 1 / xi) * log(z))
  if (!is.null(Dg) && lambda > 0) nll <- nll + lambda * sum((Dg %*% gam)^2)
  nll
}
fit_gp_basis <- function(y, Xg, lambda = 0, Dg = NULL) {
  init <- c(log(mean(y)), rep(0, ncol(Xg) - 1), 0.05)
  o <- optim(init, gp_nll, y = y, Xg = Xg, lambda = lambda, Dg = Dg,
             method = "BFGS", control = list(maxit = 600))
  list(gam = o$par[1:ncol(Xg)], xi = o$par[ncol(Xg) + 1], nll = o$value)
}

fit_gp_own <- function(x, ladder) {
  u <- ladder$U[, K]; exc <- x > u
  y <- (x - u)[exc]
  Xg  <- cbind(Btrend, H)[exc, , drop = FALSE]
  fe <- folds[exc]
  ## move one: harmonics in both, trend column the difference
  f0 <- fit_gp_basis(y, cbind(1, H[exc, , drop = FALSE]))
  f1 <- fit_gp_basis(y, cbind(1, tt[exc], H[exc, , drop = FALSE]))
  LRT <- 2 * (f0$nll - f1$nll)
  cvs <- sapply(lam_grid, function(lam) {
    mean(sapply(sort(unique(fe)), function(k) {
      tr <- fe != k
      f <- fit_gp_basis(y[tr], Xg[tr, , drop = FALSE], lam, Dtr)
      gp_nll(c(f$gam, f$xi), y[!tr], Xg[!tr, , drop = FALSE]) / sum(!tr)
    }))
  })
  lam_s <- lam_grid[which.min(cvs)]
  fs <- fit_gp_basis(y, Xg, lam_s, Dtr)
  sig <- exp(as.vector(cbind(Btrend, H) %*% fs$gam))
  list(exc = exc, u = u, n_exc = sum(exc), const = f0, lin = f1,
       LRT = LRT, p = 1 - pchisq(LRT, 1), lam_s = lam_s,
       sig = sig, xi = fs$xi, route = "own")
}

fit_gp_evgam <- function(x, ladder) {
  tryCatch({
    u <- ladder$U[, K]; exc <- x > u
    d <- data.frame(y = (x - u)[exc], tt = tt[exc], doy = doy[exc])
    f <- evgam::evgam(list(y ~ s(doy, bs = "cc", k = 8) + s(tt, k = 15), ~ 1),
                      data = d, family = "gpd",
                      knots = list(doy = c(0.5, 366.5)))
    pr <- predict(f, newdata = data.frame(tt = tt, doy = doy),
                  type = "response")
    sig <- as.numeric(pr[[1]]); xi <- as.numeric(pr[[2]][1])
    own <- fit_gp_own(x, ladder)
    gapc <- max(abs(sig - own$sig) / own$sig)
    cat(sprintf("evgam vs project sigma(t): max relative gap %.3f (REML vs
    CV smoothing; reported, not asserted)\n", gapc))
    out <- own
    out$sig <- sig; out$xi <- xi; out$route <- "evgam"
    out
  }, error = function(e) {
    cat("NOTE: evgam branch failed (", conditionMessage(e),
        ") - project fallback used\n", sep = "")
    fit_gp_own(x, ladder)
  })
}
fit_gp_stage <- function(x, ladder)
  if (has["evgam"]) fit_gp_evgam(x, ladder) else fit_gp_own(x, ladder)
```

```{r gpfit}
gC <- fit_gp_stage(xC, mC); gX <- fit_gp_stage(xX, mX); gB <- fit_gp_stage(xB, mB)
knitr::kable(data.frame(
  series = c("NTHMB sf", "BRIDG sf", "BRIDG bf"),
  route = c(gC$route, gX$route, gB$route),
  n_exc = c(gC$n_exc, gX$n_exc, gB$n_exc),
  xi = round(c(gC$xi, gX$xi, gB$xi), 3),
  LRT_trend = round(c(gC$LRT, gX$LRT, gB$LRT), 2),
  p_value = signif(c(gC$p, gX$p, gB$p), 2)),
  caption = "GP stage: route actually used, exceedance count, the constant
  shape xi (the number to quote), and move one for the trend in log sigma
  (LRT vs chi-squared(1), harmonics in both candidates).")

## sigma(t) in chunks -- the meeting's 'look at the scale in smaller pieces'
sig_tab <- round(rbind(
  `NTHMB sf` = tapply(gC$sig, list(era), mean),
  `BRIDG sf` = tapply(gX$sig, list(era), mean),
  `BRIDG bf` = tapply(gB$sig, list(era), mean)), 2)
knitr::kable(sig_tab, caption = "Mean fitted GP scale sigma(t) by era: the
chunked view of the scale curve. Read along a row for the trend in typical
overshoot size.")
```

```{r gpseason, fig.height=4.2}
op <- par(mfrow = c(1, 2), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
excd <- xX > mX$U[, K]
boxplot((xX - mX$U[, K])[excd] ~ months[excd], col = "grey92",
        xlab = "month", ylab = "raw excess",
        main = "BRIDG raw excesses by month")
boxplot(((xX - mX$U[, K]) / gX$sig)[excd] ~ months[excd], col = "grey92",
        xlab = "month", ylab = "excess / sigma(t)",
        main = "BRIDG standardized excesses by month")
par(op)
```

**Reading the figure.** Left: the raw overshoots above the threshold, by
calendar month of occurrence --- with a seasonal threshold there are
exceedance days in every month, and their raw sizes may legitimately differ
by season (freshet overshoots are physically larger). Right: the same
excesses divided by the fitted seasonal $\sigma(t)$ --- if the scale model
has absorbed the seasonality, these boxes should be level across months
with no drift; a remaining monthly pattern here is the meeting's "excesses
still seasonal" failure, localized to the GP stage rather than the
threshold.

## Per-season GP shape probe

```{r gpseasonshape}
gp_seas <- do.call(rbind, lapply(c("DJF", "MAM", "JJA", "SON"), function(sn) {
  e <- excd & seas == sn
  y <- ((xX - mX$U[, K]) / gX$sig)[e]
  f <- fit_gp_basis(y, matrix(1, sum(e), 1))
  data.frame(season = sn, n = sum(e), sigma_std = round(exp(f$gam[1]), 3),
             xi = round(f$xi, 3))
}))
knitr::kable(gp_seas, row.names = FALSE, caption = sprintf("Per-season GP on
sigma(t)-standardized BRIDG excesses. If the seasonal scale model is right,
sigma_std sits near 1 in every row. The xi column probes the constant-shape
decision (pooled xi = %.3f): a DJF value far above the others is the heavy
winter tail seen in the Exp(1) panels, and the case to discuss before ever
freeing xi.", gX$xi))
```

# Stage 1c: transform to Laplace, and the gates (v8 \S3, eqs. 5--6, pp. 3--4)

The two fitted pieces assemble into $F(x;t)$ --- ladder interpolation in
the body, GP tail above $u(t;\tau^{*})$ (v8 eq. 5, p. 3) --- and each day's
value maps through its own day's $F$ to the standard Laplace scale (v8 eq.
6, p. 4; Keef et al. 2013 eq. 2.1). Below the bottom rung ($\tau = 0.01$)
the lower tail decays exponentially, matched continuously to the bottom
ladder segment; the numerical floor sits at $1/(4N)$ and the count of days
reaching it is printed --- the old transform's hard clamp at $10^{-5}$,
which collapsed the lowest days onto the single value $-2.44$ and drew the
flat-then-jump QQ artifact, is gone, and this instrumentation proves it on
every knit.

```{r transform}
qlap <- function(p) ifelse(p < 0.5, log(2 * p), -log(2 * (1 - p)))
plap <- function(y) ifelse(y < 0, exp(y) / 2, 1 - exp(-y) / 2)

F_of <- function(x, ladder, gp) {
  U <- ladder$U
  Fv <- rep(NA_real_, length(x))
  above <- x > U[, K]
  S <- (1 + gp$xi * (x - U[, K]) / gp$sig)
  Fv[above] <- 1 - (1 - params$tau_star) * pmax(S[above], 0)^(-1 / gp$xi)
  for (k in seq_len(K - 1)) {
    m <- !above & x > U[, k] & x <= U[, k + 1]
    Fv[m] <- taus[k] + (taus[k + 1] - taus[k]) *
             (x[m] - U[m, k]) / (U[m, k + 1] - U[m, k])
  }
  low <- !above & x <= U[, 1]
  beta1 <- (U[, 2] - U[, 1]) / log(taus[2] / taus[1])   # continuous match
  Fv[low] <- taus[1] * exp((x[low] - U[low, 1]) / beta1[low])
  n_low <- sum(low); n_floor <- sum(Fv[low] <= 1 / (4 * N))
  cat(sprintf("  below bottom rung: %d days (%.2f%%); at numerical floor: %d\n",
      n_low, 100 * n_low / length(x), n_floor))
  pmin(pmax(Fv, 1 / (4 * N)), 1 - 1e-9)
}
cat("NTHMB sf:\n"); FC <- F_of(xC, mC, gC)
cat("BRIDG sf:\n"); FX <- F_of(xX, mX, gX)
cat("BRIDG bf:\n"); FB <- F_of(xB, mB, gB)
YC <- qlap(FC); YX <- qlap(FX); YB <- qlap(FB)
y_star <- -log(2 * (1 - params$tau_star))    # 2.303 at 0.95 (v8 Sec 3, p. 4)
```

```{r pitgate, fig.height=4}
op <- par(mfrow = c(1, 3), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
for (nm in c("FC", "FX", "FB")) {
  hist(get(nm), breaks = 40, col = "grey88", border = "white", freq = FALSE,
       xlab = "fitted F(x; t)", main = paste("gate:",
       c(FC = "NTHMB sf", FX = "BRIDG sf", FB = "BRIDG bf")[nm]))
  abline(h = 1, col = "firebrick3", lwd = 2)
}
par(op)
```

**Reading the figure.** The gate of stage 1 (v8 \S3, p. 4). The x-axis is
each day's fitted probability under its own day's distribution; the y-axis
density. If the marginal model is right these are Unif(0,1): flat bars on
the red line. A spike in the first bar was the old clamp's signature; a
bump near 1 says the tail is too light; waves in the body point at the
ladder; and --- the meeting's verdict on the previous version --- a shape
that reappears when sliced by season is the seasonality the margins missed,
which the seasonal basis exists to remove. Downstream stages wait here.

## Attributing the histogram: rung audit, per-bar noise band, residual maps

```{r rungaudit}
rung_audit <- function(x, ladder)
  sapply(seq_len(K), function(k) mean(x <= ladder$U[, k]) - taus[k])
ra <- round(rbind(`NTHMB sf` = rung_audit(xC, mC),
                  `BRIDG sf` = rung_audit(xX, mX),
                  `BRIDG bf` = rung_audit(xB, mB)), 4)
colnames(ra) <- paste0("tau=", taus)
knitr::kable(ra, caption = "Rung calibration audit: share of days at or
below each fitted rung, minus its target tau. Zero everywhere = a perfectly
calibrated ladder; a negative entry at tau = 0.01 means the bottom rung sits
too low (the short first histogram bar); an interior entry off zero places a
histogram wave at that rung of the qgam fit.")

ra_era <- sapply(seq_len(K), function(k)
  tapply(xX <= mX$U[, k], era, mean) - taus[k])
colnames(ra_era) <- paste0("tau=", taus)
knitr::kable(round(ra_era, 3), caption = "BRIDG sf: the same audit per era
-- an entry changing sign along a row is drift the additive seasonal basis
missed (the te case), not a global rung offset.")
```

```{r pitdev, fig.width=9.5, fig.height=4}
brks <- seq(0, 1, by = 0.025); mids <- brks[-1] - 0.0125
for (nm in c("FC", "FX", "FB")) {
  Fv <- get(nm); lab <- c(FC = "NTHMB sf", FX = "BRIDG sf", FB = "BRIDG bf")[nm]
  dens <- hist(Fv, breaks = brks, plot = FALSE)$density
  bootd <- replicate(200, {
    ky <- years %in% sample(unique(years), replace = TRUE)
    hist(Fv[ky], breaks = brks, plot = FALSE)$density
  })
  bb <- apply(bootd, 1, quantile, c(0.025, 0.975))
  bad <- which(bb[1, ] > 1 | bb[2, ] < 1)
  plot(mids, dens, type = "h", lwd = 5, col = "grey70",
       ylim = range(bb, dens, 1), xlab = "fitted F(x; t)", ylab = "density",
       main = paste(lab, "- bars whose year-block band excludes 1 are red"))
  segments(mids, bb[1, ], mids, bb[2, ], col = adjustcolor("steelblue3", 0.6))
  abline(h = 1, col = "firebrick3", lwd = 2)
  points(mids[bad], dens[bad], pch = 16, col = "firebrick3")
  cat(lab, "- genuinely off bins at:",
      paste(sprintf("%.3f", mids[bad]), collapse = " "), "\n")
}
```

**Reading the figure.** Each blue segment answers: how much would this bar
move if history had dealt a different set of years? Two hundred re-draws of
whole years (the block that preserves clusters and seasons), the 2.5--97.5%
range per bar; a bar is flagged red only when its whole interval misses 1
--- structural, not year-to-year luck. The fit is not redone per replicate,
so this measures sampling noise, not fit uncertainty.

```{r pitwho, fig.width=9.5, fig.height=4.5}
comp_map <- function(Fv, fac, main) {
  nb <- length(brks) - 1
  Ocnt <- table(fac, cut(Fv, brks, include.lowest = TRUE))
  Ecnt <- rowSums(Ocnt) %o% rep(1 / nb, nb)
  R <- (Ocnt - Ecnt) / sqrt(Ecnt)
  image(1:nb, 1:nrow(R), t(R[nrow(R):1, ]), zlim = c(-6, 6),
        col = hcl.colors(25, "Blue-Red 2"), xlab = "PIT bin (low to high F)",
        ylab = "", yaxt = "n", main = main)
  axis(2, at = 1:nrow(R), labels = rev(rownames(R)), las = 2, cex.axis = 0.7)
  wr <- which(abs(R) == max(abs(R)), arr.ind = TRUE)[1, ]
  cat(main, "| largest cell:", rownames(R)[wr[1]],
      sprintf("bin (%.3f, %.3f], residual %+.1f\n",
              brks[wr[2]], brks[wr[2] + 1], R[wr[1], wr[2]]))
}
mon <- factor(month.abb[months], levels = month.abb)
comp_map(FC, mon, "NTHMB sf: month x PIT bin (red = too many days)")
comp_map(FC, era, "NTHMB sf: era x PIT bin")
comp_map(FX, mon, "BRIDG sf: month x PIT bin")
comp_map(FX, era, "BRIDG sf: era x PIT bin")
```

**Reading the maps.** Cells are Pearson residuals (O - E)/sqrt(E) of a
month-by-bin (or era-by-bin) table under within-slice uniformity: |R| <= 2
is noise, |R| >= 4 is structure; look for stripes, not lone cells. A month
row stripe = residual seasonality (the forbidden outcome); a bin column
stripe = one rung miscalibrated for all months (ladder tuning); an era row
stripe = drift the additive basis missed (the te case).

## Pooled and sliced Laplace QQ (v8 \S3, p. 4)

```{r slicedqq, fig.height=7}
qq_laplace <- function(y, main) {
  n <- length(y); th <- qlap((1:n) / (n + 1))
  plot(th, sort(y), pch = 16, cex = 0.35, col = "steelblue4",
       xlab = "Laplace quantiles", ylab = "sample", main = main,
       cex.main = 0.9)
  abline(0, 1, col = "grey55"); abline(h = y_star, v = y_star,
                                       col = "grey85", lty = 3)
}
dec1 <- min(years):(min(years) + 9)
dec2 <- (max(years) - 9):max(years)
op <- par(mfrow = c(2, 5), mar = c(3, 3, 2, 0.6), mgp = c(1.9, 0.6, 0))
for (yv in list(c("YC", "NTHMB"), c("YX", "BRIDG"))) {
  Y <- get(yv[1])
  qq_laplace(Y, paste(yv[2], "pooled"))
  qq_laplace(Y[seas == "DJF"], paste(yv[2], "DJF"))
  qq_laplace(Y[seas == "JJA"], paste(yv[2], "JJA"))
  qq_laplace(Y[years %in% dec1], paste(yv[2], min(dec1), "-", max(dec1)))
  qq_laplace(Y[years %in% dec2], paste(yv[2], min(dec2), "-", max(dec2)))
}
par(op)
```

**Reading the figure.** Rows are stations; columns are the pooled sample,
two season slices, and the first and last decades. Each panel plots sorted
Laplace-scale values against standard Laplace quantiles; the grey diagonal
is perfect agreement and the dotted lines mark $y^{*}$, where the
dependence stage begins. The pooled panel's body is near-straight almost by
construction (the ladder interpolation tracks the data); its information is
the corner beyond $y^{*}$, where the GP does real work --- and the lower
tail, where the old clamp's flat-run-then-jump would sit: a remaining short
flat segment there is genuine ties in the data (Section 2's forensics name
them), not the transform. The season panels are the meeting's test: with
the seasonal margins they must hug the line --- a summer slice shifted
bodily off the diagonal was the trend-only design's signature and must be
gone. Decade panels off the line say the slow curves missed the drift.

```{r excexp, fig.height=4}
op <- par(mfrow = c(1, 4), mar = c(3, 3, 2, 0.6), mgp = c(1.9, 0.6, 0))
for (sn in c("DJF", "MAM", "JJA", "SON")) {
  e <- YC[seas == sn & YC > y_star] - y_star
  if (length(e) > 10) {
    n <- length(e); th <- -log(1 - (1:n) / (n + 1))
    plot(th, sort(e), pch = 16, cex = 0.4, col = "steelblue4",
         xlab = "Exp(1) quantiles", ylab = "excess above y*",
         main = paste("NTHMB", sn, "(n =", n, ")"), cex.main = 0.9)
    abline(0, 1, col = "grey55")
  } else plot.new()
}
par(op)
```

**Reading the figure.** The decisive tail check (v8 \S3, p. 4): on the
Laplace scale the excesses $Y - y^{*}$ above the dependence threshold are
exactly unit exponential if the transform is honest, so each season's
exceedances are QQ-plotted against Exp(1). These are the very days the
dependence stage consumes --- if these panels pass, stage 2 stands on solid
ground even where the deep winter body is imperfect. Panel headers carry
the exceedance count per season: with a seasonal threshold every season
contributes days, and a season with almost none is itself information.

```{r decshare}
dec_all <- (years - min(years)) %/% 10
share <- t(sapply(list(NTHMB = YC, BRIDG = YX), function(Y)
  round(tapply(Y > y_star, dec_all, mean), 3)))
colnames(share) <- paste0(min(years) + 10 * as.integer(colnames(share)), "s")
knitr::kable(share, caption = sprintf("Share of days above y* = %.3f per
decade; target %.2f in every cell (v8 Sec 3, p. 4 -- the moving threshold's
purpose, LL20).", y_star, 1 - params$tau_star))
```

# Stage 2: the dependence model (v8 \S5, eq. 8, p. 11) --- project code

No package fits the non-stationary Heffernan--Tawn model, so this fitter is
the project's one self-implemented core. On the Laplace scale $a(t) \in
[-1, 1]$ (link $a = 2\,\mathrm{plogis}(\eta) - 1$) and $b(t) < 1$; the
working likelihood stays Gaussian --- Keef et al. tested Normal against
Laplace working residuals and kept the Normal (v8 \S5, p. 13) --- and the
maximization carries the Keef ordering constraint (v8 eq. 9, p. 13): beyond
the data the fitted conditional quantiles may not exceed what perfect
positive dependence would give, nor undercut perfect negative dependence.
The constraint is imposed at $v_K$ above the largest observed conditioning
value (v8 \S3, p. 4), checked at the residual range endpoints, and --- our
flagged extension --- pointwise in $t$ for curve models. Verification is
two-tier: the all-constant case against texmex *on its defaults* (Laplace
margins, Keef constraints --- the reference implementation, v8 \S4.5, p. 10),
and the fixed-$b$ linear case against `mgcv::gam(gaulss)`.

```{r winchi, fig.width=9.5, fig.height=4.5}
pobs <- function(x) rank(x, ties.method = "average") / (length(x) + 1)
win_chi <- function(c1, c2, y0, y1, u = 0.95) {
  k <- years >= y0 & years <= y1
  uu <- pobs(c1[k]); vv <- pobs(c2[k])
  sum(uu > u & vv > u) / (sum(k) * (1 - u))
}
w_starts <- seq(min(years), max(years) - 5, by = 1)
chi_seq <- sapply(w_starts, function(y0) win_chi(xC, xX, y0, y0 + 5))
plot(w_starts + 2.5, chi_seq, pch = 16, cex = 0.6, col = "steelblue4",
     ylim = c(0, max(chi_seq) * 1.1), xlab = "window centre (year)",
     ylab = "chi(0.95), 6-year window",
     main = "Model-free screening: windowed chi, BRIDG-NTHMB, this file")
lines(lowess(w_starts + 2.5, chi_seq, f = 0.3), col = "firebrick3", lwd = 2)
abline(h = 0.05, lty = 2, col = "grey55")
cat(sprintf("windowed chi: mean of first 10 windows %.3f | last 10 %.3f\n",
    mean(head(chi_seq, 10)), mean(tail(chi_seq, 10))))
```

**Reading the figure.** Each dot is one six-year window: the share of days
on which both stations exceed their own pooled 0.95 rank level, divided by
5% --- the empirical chi(0.95), computed from ranks alone, no model. The
dashed line marks independence; near 1 is asymptotic dependence; the red
curve is the smoothed path. This is the yardstick every fitted dependence
result must face: if the fitted a(t) is flat while this curve climbs, the
fit is losing a real signal.

```{r htmachinery}
v_dep <- -log(2 * (1 - params$kappa))
keep <- YC > v_dep
yCe <- YC[keep]; yXe <- YX[keep]; tte <- tt[keep]
fe <- folds[keep]; ye <- years[keep]
n_e <- sum(keep); cat("exceedance days above v =", round(v_dep, 3), ":", n_e, "\n")

v_K <- max(yCe) + 0.5
zp <- range(yXe - yCe); zm <- range(yXe + yCe)   # Keef bound residual ranges

Bdep <- make_basis(tt, 2 * params$spacing_yr, span_yr)
Bde <- Bdep[keep, , drop = FALSE]
p_dep <- ncol(Bdep)
tdec <- quantile(tte, 0:10 / 10)                  # constraint checked here

keef_on <- TRUE
keef_viol <- function(a_v, b_v, Wlo, Whi) {
  if (!keef_on) return(FALSE)
  ## ordering (v8 eq. 9, p. 13) at y in {v_K, v_K + 3}, q in {0, 1},
  ## for each (a, b) supplied (deciles of t for curve models)
  for (y in c(v_K, v_K + 3)) {
    lo <- a_v * y + y^b_v * Wlo; hi <- a_v * y + y^b_v * Whi
    if (any(hi > y + zp[2]) || any(lo < -y + zm[1])) return(TRUE)
  }
  FALSE
}

ht_nll <- function(theta, yC, yX, Bmat, lambda = 0, w = c(1, 1, 10, 10),
                   Bchk = NULL) {
  p <- ncol(Bmat)
  ga <- theta[1:p]; gb <- theta[(p+1):(2*p)]
  gm <- theta[(2*p+1):(3*p)]; gs <- theta[(3*p+1):(4*p)]
  a <- 2 * plogis(as.vector(Bmat %*% ga)) - 1
  b <- as.vector(Bmat %*% gb)
  m <- as.vector(Bmat %*% gm); s <- exp(as.vector(Bmat %*% gs))
  if (any(b > 0.999)) return(1e10)
  loc <- a * yC + m * yC^b; sc <- s * yC^b
  if (any(!is.finite(sc)) || any(sc <= 0)) return(1e10)
  W <- (yX - a * yC) / yC^b
  if (is.null(Bchk)) Bchk <- Bmat[c(1, nrow(Bmat)), , drop = FALSE]
  a_chk <- 2 * plogis(as.vector(Bchk %*% ga)) - 1
  b_chk <- as.vector(Bchk %*% gb)
  if (keef_viol(a_chk, b_chk, min(W), max(W))) return(1e10)
  sum(log(sc) + 0.5 * ((yX - loc) / sc)^2) +
    lambda * (w[1] * sum(diff(ga)^2) + w[2] * sum(diff(gb)^2) +
              w[3] * sum(diff(gm)^2) + w[4] * sum(diff(gs)^2))
}
fit_ht_basis <- function(yC, yX, Bmat, lambda = 0, init = NULL, Bchk = NULL) {
  p <- ncol(Bmat)
  if (is.null(init)) init <- c(rep(0, p), rep(0.2, p), rep(0, p), rep(0, p))
  optim(init, ht_nll, yC = yC, yX = yX, Bmat = Bmat, lambda = lambda,
        Bchk = Bchk, method = "BFGS", control = list(maxit = 800))
}
curves_of <- function(theta, Bmat) {
  p <- ncol(Bmat)
  list(a = 2 * plogis(as.vector(Bmat %*% theta[1:p])) - 1,
       b = as.vector(Bmat %*% theta[(p+1):(2*p)]),
       m = as.vector(Bmat %*% theta[(2*p+1):(3*p)]),
       s = exp(as.vector(Bmat %*% theta[(3*p+1):(4*p)])))
}
```

```{r htfit}
B1e <- matrix(1, n_e, 1); B2e <- cbind(1, tte)

## texmex on its own defaults (Laplace margins + Keef constraints) is the
## reference: it sees the already-Laplace pair, so its own marginal step is
## near-identity and tier 1 isolates the dependence fitter
init0 <- NULL
if (has["texmex"]) init0 <- tryCatch({
  tx <- texmex::mex(data.frame(C = YC, X = YX), mqu = params$tau_star,
                    dqu = params$kappa, which = "C")
  co <- tx$dependence$coefficients
  c(qlogis(min(max((co[1, 1] + 1) / 2, 1e-3), 1 - 1e-3)), co[2, 1], 0, 0)
}, error = function(e) NULL)

h0 <- fit_ht_basis(yCe, yXe, B1e, init = init0)
h1 <- fit_ht_basis(yCe, yXe, B2e, init = as.vector(rbind(h0$par, 0)),
                   Bchk = cbind(1, tdec))
LRT_dep <- 2 * (h0$value - h1$value)   # 4 df, working likelihood: guide only

lam_grid_dep <- 10^(0:6)
Bchk_dep <- predict(Bdep, tdec)
cv_dep <- sapply(lam_grid_dep, function(lam) {
  mean(sapply(sort(unique(fe)), function(k) {
    tr <- fe != k
    f <- fit_ht_basis(yCe[tr], yXe[tr], Bde[tr, , drop = FALSE], lam,
                      Bchk = Bchk_dep)
    ht_nll(f$par, yCe[!tr], yXe[!tr], Bde[!tr, , drop = FALSE], 0,
           Bchk = Bchk_dep) / sum(!tr)
  }))
})
lam_dep <- lam_grid_dep[which.min(cv_dep)]
hs <- fit_ht_basis(yCe, yXe, Bde, lam_dep,
        init = c(rep(h0$par[1], p_dep), rep(h0$par[2], p_dep),
                 rep(h0$par[3], p_dep), rep(h0$par[4], p_dep)),
        Bchk = Bchk_dep)

## model selection, printed: the spline is kept only if it beats the line
## at its own game (a worse penalized fit = collapse or the a-m ridge)
spline_ok <- hs$value < h1$value
sel <- if (spline_ok) list(par = hs$par, B = Bdep, label = "spline") else
                      list(par = h1$par, B = cbind(1, tt), label = "linear")
cs_sel <- curves_of(sel$par, sel$B)
cat("selected dependence model:", sel$label, "\n")

knitr::kable(data.frame(
  model = c("all constant", "all linear", "spline (penalized)"),
  working_nll = round(c(h0$value, h1$value, hs$value), 1),
  note = c(sprintf("a = %.3f, b = %.3f%s", 2 * plogis(h0$par[1]) - 1,
                   h0$par[2], ifelse(is.null(init0), "", " (texmex-seeded)")),
           sprintf("LRT(4df) = %.2f -- approximate guide; the bootstrap
                    interval is the honest test (v8 Sec 4.6, p. 10)", LRT_dep),
           sprintf("lambda_dep = %g by blocked CV; %s", lam_dep,
                   ifelse(spline_ok, "kept", "rejected (collapse) -- linear
                   selected for all downstream output")))),
  caption = "Dependence stage, Case 4 (BRIDG | NTHMB), v8 Sec 5 eq. (8)
  p. 11, Laplace scale, Keef constraints active.")
```

## Dependence by season, and the freshet-only refit

```{r seasondep}
sdep <- do.call(rbind, lapply(c("DJF", "MAM", "JJA", "SON"), function(sn) {
  k <- keep & seas == sn
  f <- fit_ht_basis(YC[k], YX[k], matrix(1, sum(k), 1))
  data.frame(season = sn, n = sum(k),
             a_hat = round(2 * plogis(f$par[1]) - 1, 3),
             b_hat = round(f$par[2], 3))
}))
knitr::kable(sdep, row.names = FALSE, caption = "Constant HT fit per season
of the conditioning day. A high-freshet / low-winter pattern says a flat
pooled a(t) is dilution by the season-relative conditioning event -- the
seasonal threshold makes 'NTHMB extreme' include winter days whose
co-movement with BRIDG is weak -- not absence of dependence.")

kF <- keep & months %in% 5:8
h0F <- fit_ht_basis(YC[kF], YX[kF], matrix(1, sum(kF), 1))
h1F <- fit_ht_basis(YC[kF], YX[kF], cbind(1, tt[kF]),
                    init = as.vector(rbind(h0F$par, 0)), Bchk = cbind(1, tdec))
cat(sprintf("freshet-only (May-Aug, n = %d): constant a = %.3f, b = %.3f |
  linear LRT(4df) = %.2f, a(t) endpoints %.3f -> %.3f\n",
  sum(kF), 2 * plogis(h0F$par[1]) - 1, h0F$par[2],
  2 * (h0F$value - h1F$value),
  2 * plogis(h1F$par[1]) - 1, 2 * plogis(h1F$par[1] + h1F$par[2]) - 1))
```

## The constraint-off sanity refit

```{r keeffree}
keef_on <- FALSE
h1_free <- fit_ht_basis(yCe, yXe, B2e, init = as.vector(rbind(h0$par, 0)))
keef_on <- TRUE
cat(sprintf("linear fit, Keef wall OFF: nll %.1f (constrained %.1f) |
  a(t) endpoints %.3f -> %.3f (constrained: %.3f -> %.3f)\n",
  h1_free$value, h1$value,
  2 * plogis(h1_free$par[1]) - 1,
  2 * plogis(h1_free$par[1] + h1_free$par[2]) - 1,
  2 * plogis(h1$par[1]) - 1, 2 * plogis(h1$par[1] + h1$par[2]) - 1))
```

If the free fit lands on the same objective and the same endpoints, the
zero slope is genuine; if it finds a lower objective and a real slope, the
constraint's hard 1e10 wall stalled BFGS (the KNOWN DEFECT of the Aug-22
run: 2698.6 vs 2022.2, endpoints 0.095/0.095 vs 0.396/0.490). The repair --
smooth penalty in place of the wall, Nelder-Mead then BFGS, multiple starts
-- is fix-list item 1 in walkthrough_v4.

## Tier 1: texmex on its defaults (all-constant case)

```{r tier1, eval=has[["texmex"]] && exists("tx")}
co <- tx$dependence$coefficients
knitr::kable(data.frame(
  parameter = c("a", "b"),
  project = round(c(2 * plogis(h0$par[1]) - 1, h0$par[2]), 4),
  texmex = round(co[1:2, 1], 4)),
  caption = "Tier 1: all-constant project fit vs texmex on its own defaults
  (Laplace margins, Keef constraints -- the reference implementation, v8
  Sec 4.5 p. 10), both on the SAME Laplace-scale pair so the margins are
  shared and the gap isolates the fitter. The gap should be small; a large
  one is a fitter problem to investigate, not to explain away. (The Keef
  implementations differ in detail: texmex imposes Theorem 1 exactly, the
  project fitter checks the ordering at the residual range endpoints.)")
```

## Tier 2: the mgcv check (fixed $b$, linear curves)

With $b$ frozen at $\hat b$ and the remaining curves linear, the model is a
varying-coefficient Gaussian location-scale model --- mean $= a(t)y +
m(t)y^{b}$, log sd $= \log s(t) + b \log y$ --- which `mgcv::gam(gaulss)`
fits by ML. Same model, two optimizers; the fitted means must agree. To
keep it literally the same model, the check fitter uses an identity link
for $a(t)$ (mgcv's varying coefficient is linear in $t$); the production
fitter's bounded link and Keef constraint are deliberately not part of
what this tier tests.

```{r tier2, eval=has[["mgcv"]]}
b0 <- h0$par[2]
## identity link for a(t) HERE ONLY: tier 2 is same-model-two-optimizers,
## and mgcv's varying coefficient is linear in t, so the check model must
## be too -- the production fitter's bounded link is not what is tested
ht_nll_fixedb <- function(th, yC, yX, ttv, b0) {
  a <- th[1] + th[2] * ttv
  m <- th[3] + th[4] * ttv; s <- exp(th[5] + th[6] * ttv)
  loc <- a * yC + m * yC^b0; sc <- s * yC^b0
  sum(log(sc) + 0.5 * ((yX - loc) / sc)^2)
}
thf <- optim(c(2 * plogis(h0$par[1]) - 1, 0, h0$par[3], 0, h0$par[4], 0),
             ht_nll_fixedb, yC = yCe, yX = yXe, ttv = tte, b0 = b0,
             method = "BFGS", control = list(maxit = 2000))
a_f <- thf$par[1] + thf$par[2] * tte
loc_own <- a_f * yCe + (thf$par[3] + thf$par[4] * tte) * yCe^b0
sc_own  <- exp(thf$par[5] + thf$par[6] * tte) * yCe^b0

yCb <- yCe^b0
gmfit <- mgcv::gam(list(yXe ~ -1 + yCe + I(tte * yCe) + yCb + I(tte * yCb),
                        ~ tte + offset(b0 * log(yCe))),
                   family = mgcv::gaulss())
lp <- predict(gmfit)
loc_m <- lp[, 1]
sd_a <- exp(lp[, 2]); sd_b <- 1 / (exp(lp[, 2]) + 0.01)
pick <- which.min(abs(c(sd((yXe - loc_m) / sd_a),
                        sd((yXe - loc_m) / sd_b)) - 1))
sc_m <- if (pick == 1) sd_a else sd_b

t2 <- c(loc_rel = max(abs(loc_own - loc_m)) / diff(range(yXe)),
        sc_rel = median(abs(sc_own - sc_m) / sc_own))
stopifnot(t2["loc_rel"] < 0.02)
knitr::kable(data.frame(check = c("conditional mean (max gap, range units)",
                                  "conditional sd (median relative gap)"),
                        value = signif(t2, 3)),
             caption = "Tier 2: project fitter (b fixed, linear curves,
             unconstrained) vs mgcv::gam(gaulss) on the identical model.
             Mean structure asserted at 2%; sd reported (link-convention
             sensitive). Note the mgcv fit carries no Keef constraint, so
             the project fitter runs this case unconstrained too -- the
             comparison is only clean like for like.")
```

Note for the tier-2 chunk: the fixed-$b$ comparison model is fitted without
the Keef check (mgcv cannot impose it), so if the main fit sat on the
constraint boundary the two would legitimately differ --- the printed
constant fit's distance from the boundary says whether that caveat bites.

## Diagnostics and the anchor (v8 \S5, pp. 11--12)

```{r htdiag, fig.height=4.2}
a_e <- if (sel$label == "spline") curves_of(sel$par, Bde) else
       curves_of(sel$par, cbind(1, tte))
Zhat <- ((yXe - a_e$a * yCe) / yCe^a_e$b - a_e$m) / a_e$s
op <- par(mfrow = c(1, 2), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
plot(yCe, Zhat, pch = 16, cex = 0.4, col = adjustcolor("steelblue4", 0.5),
     xlab = expression(Y[C]), ylab = expression(hat(Z)),
     main = "residual vs conditioning value")
lines(lowess(yCe, Zhat, f = 0.5), col = "firebrick3", lwd = 2)
era_e <- cut(ye, seq(min(years), max(years) + 20, by = 20), right = FALSE)
boxplot(Zhat ~ era_e, col = "grey92", las = 2, cex.axis = 0.7,
        xlab = "", ylab = expression(hat(Z)),
        main = "residual by era (covariate-free G, v8 Sec 5 p. 11)")
abline(h = 0, lty = 3, col = "grey60")
par(op)
Ztable <- Zhat
```

**Reading the figure.** Left: x-axis the conditioning value $Y_C$ (Laplace
scale, above $v$), y-axis the standardized residual from the *selected*
model; the red line is a running mean. Flat at zero with constant scatter
says the fitted curves absorbed everything systematic; drift in level or
spread says the normalization failed (v8 \S5, p. 11). Right: the residuals
by era --- equal boxes are the covariate-free-$G$ assumption, the licence
for reusing any era's residual at any other $t$ (v8 \S5, p. 12).

```{r anchor}
mA_C <- fit_ladder_cached(get_series(A_C, "sf"), "AC"); gA_C <- fit_gp_stage(get_series(A_C, "sf"), mA_C)
mA_X <- fit_ladder_cached(get_series(A_X, "sf"), "AX"); gA_X <- fit_gp_stage(get_series(A_X, "sf"), mA_X)
YA_C <- qlap(F_of(get_series(A_C, "sf"), mA_C, gA_C))
YA_X <- qlap(F_of(get_series(A_X, "sf"), mA_X, gA_X))
ka <- YA_C > v_dep
h_anchor <- fit_ht_basis(YA_C[ka], YA_X[ka], matrix(1, sum(ka), 1))
cat(sprintf("Case 1 anchor: a-hat = %.3f (must sit near 1; v8 Sec 5 p. 12),
            b-hat = %.3f\n", 2 * plogis(h_anchor$par[1]) - 1, h_anchor$par[2]))
```

# The conditional-simulation display (meeting 62, rebuilt)

Two decisions from the meeting shape this section. *Condition on the day of
year*: a randomly picked date lands in low-flow seasons and the simulated
points come out smaller than the real extremes; the display day is now the
climatological freshet peak (the day of year at which the fitted threshold
is highest), a parameter of this chunk. *No year colours on the standard
scale*: after the marginal transform every day is the same Laplace
distribution, so the standard-margins panel shows one model cloud against
the data; the year contrast appears only on the original scale, where the
margins genuinely differ by year --- and the scenario contrast (RCP 4.5
against 8.5) takes over as the headline comparison once the 8.5 export
joins the file. Which scale the paper's version uses is the open display
question for Friday; both are drawn here.

```{r pseudolap, fig.height=5.2}
## reverse direction fitted with the same machinery (v8 Sec 5, p. 11)
keepR <- YX > v_dep
hR <- fit_ht_basis(YX[keepR], YC[keepR], cbind(1, tt[keepR]),
                   Bchk = cbind(1, tdec))
csR <- curves_of(hR$par, cbind(1, tt[keepR]))
ZtabR <- ((YC[keepR] - csR$a * YX[keepR]) / YX[keepR]^csR$b - csR$m) / csR$s

doy_star <- as.integer(names(which.max(tapply(mX$U[, K], doy, mean))))
Yr_mid <- round(mean(range(years)))
cat("display day of year (freshet peak):", doy_star,
    "| standard-scale reference year:", Yr_mid, "\n")

gen_pseudo <- function(a, b, m, s, Ztab, n = 600) {
  yc <- qlap(params$kappa + (1 - params$kappa) * runif(n))
  z <- sample(Ztab, n, replace = TRUE)
  cbind(yc = yc, yx = a * yc + yc^b * (m + s * z))
}
at_day <- function(cv, Yr) {          # selected-model curves on one day
  i <- which(years == Yr & doy == doy_star)[1]
  if (is.na(i)) i <- which(years == Yr & doy == doy_star - 1)[1]
  list(a = cv$a[i], b = cv$b[i], m = cv$m[i], s = cv$s[i], idx = i)
}
cs_full <- curves_of(sel$par, if (sel$label == "spline") Bdep else cbind(1, tt))
csR_full <- curves_of(hR$par, cbind(1, tt))

op <- par(mfrow = c(1, 2), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
pd <- at_day(cs_full, Yr_mid)
plot(YC, YX, pch = ".", col = "grey35",
     xlab = "Y (NTHMB, Laplace)", ylab = "Y (BRIDG, Laplace)",
     main = sprintf("(a) model pseudosample of (Y_C, Y_X), year %d", Yr_mid),
     cex.main = 0.9)
abline(v = v_dep, lty = 2, col = "grey55")
ps <- gen_pseudo(pd$a, pd$b, pd$m, pd$s, Ztable)
points(ps[, 1], ps[, 2], cex = 0.45, col = adjustcolor("steelblue3", 0.6))
legend("topleft", c("model pseudosample", "data"),
       col = c("steelblue3", "grey35"), pch = c(1, 46), bty = "n", cex = 0.8)
pdR <- at_day(csR_full, Yr_mid)
plot(YX, YC, pch = ".", col = "grey35",
     xlab = "Y (BRIDG, Laplace)", ylab = "Y (NTHMB, Laplace)",
     main = sprintf("(b) reverse direction, year %d", Yr_mid),
     cex.main = 0.9)
abline(v = v_dep, lty = 2, col = "grey55")
psR <- gen_pseudo(pdR$a, pdR$b, pdR$m, pdR$s, ZtabR)
points(psR[, 1], psR[, 2], cex = 0.45, col = adjustcolor("steelblue3", 0.6))
par(op)
```

**Reading the figure.** Both axes are Laplace-scale values; panel (a)
conditions on NTHMB (x-axis), panel (b) on BRIDG. Dots are all observed
days; the dashed line is the dependence threshold $v$; circles are the
model's pseudosample to the right of $v$, generated at the freshet-peak day
of the mid-axis year. The eye's goodness-of-fit test: circles living where
dots live is agreement; circles beyond the largest dots are the
extrapolation, shown rather than asserted. One colour only: on this scale
the margins carry no year, so a year contrast here would draw a
distinction the axes cannot show (the joint drift is real but lives in the
dependence curves; it is quantified by the fitted $a(t)$ and the
functionals below, and displayed on the original scale next). Comparing
panels (a) and (b) is the self-consistency caution (v8 \S5, p. 11).

```{r pseudoorig, fig.height=5}
q_inv <- function(p, ladder, gp, idx) {
  U <- ladder$U[idx, ]; sig <- gp$sig[idx]; xi <- gp$xi
  beta1 <- (U[2] - U[1]) / log(taus[2] / taus[1])
  sapply(p, function(pp) {
    if (pp >= params$tau_star)
      U[K] + sig / xi * (((1 - params$tau_star) / (1 - pp))^xi - 1)
    else if (pp < taus[1]) U[1] + beta1 * log(pp / taus[1])
    else {
      k <- max(which(taus <= pp))
      if (k == K) U[k]
      else U[k] + (U[k + 1] - U[k]) * (pp - taus[k]) / (taus[k + 1] - taus[k])
    }
  })
}
yr_show <- c(params$start_year + 6, max(years) - 4)
cols2 <- c("steelblue3", "firebrick3")
plot(xC, xX, pch = ".", col = "grey80", log = "xy",
     xlab = "NTHMB daily flow", ylab = "BRIDG daily flow",
     main = sprintf("model pseudosamples on the original scale, day-of-year
     %d, years %d and %d", doy_star, yr_show[1], yr_show[2]), cex.main = 0.9)
for (e in 1:2) {
  pdE <- at_day(cs_full, yr_show[e])
  ps <- gen_pseudo(pdE$a, pdE$b, pdE$m, pdE$s, Ztable)
  xc_o <- q_inv(pmin(plap(ps[, 1]), 1 - 1e-9), mC, gC, pdE$idx)
  xx_o <- q_inv(pmin(pmax(plap(ps[, 2]), 1 / (4 * N)), 1 - 1e-9), mX, gX, pdE$idx)
  points(xc_o, xx_o, cex = 0.45, col = adjustcolor(cols2[e], 0.55))
}
amx <- tapply(seq_len(N), years, function(i) i[which.max(xC[i])])
points(xC[amx], xX[amx], pch = 17, cex = 0.7, col = "grey25")
legend("topleft", c(paste("pseudo", yr_show[1]), paste("pseudo", yr_show[2]),
                    "annual-max days (NTHMB)", "all days"),
       col = c(cols2, "grey25", "grey80"), pch = c(1, 1, 17, 46),
       bty = "n", cex = 0.8)
```

**Reading the figure.** The same pseudosamples pushed back through each
day's fitted margins onto the original flow scale (log axes), generated at
the freshet-peak day of an early and a late year --- here the year contrast
is meaningful, because the margins differ by year on this scale. The dark
triangles are the observed annual-maximum days of the conditioning station:
the meeting's complaint --- simulated points smaller than the real extremes
--- is the check here; with day-of-year conditioning the clouds must reach
and extend the triangles' territory. The blue-to-red movement is the whole
fitted change --- margins and dependence together; whether the paper shows
this original-scale version or the standard-scale one is the Friday
question, and the scenario contrast will replace the year contrast once
RCP 8.5 is in the file.

# Stage 3: extremal index (v8 \S6, eq. 13, p. 15)

```{r theta}
ferro_segers <- function(exc_idx) {
  Ti <- diff(exc_idx); n <- length(exc_idx)
  if (max(Ti) <= 2) min(1, 2 * sum(Ti)^2 / ((n - 1) * sum(Ti^2)))
  else min(1, 2 * sum(Ti - 1)^2 / ((n - 1) * sum((Ti - 1) * (Ti - 2))))
}
theta_route <- function(x, u) {
  own <- ferro_segers(which(x > u))
  if (has["extRemes"]) {
    pkg <- tryCatch(as.numeric(
      extRemes::extremalindex(x - u, threshold = 0, method = "intervals")[1]),
      error = function(e) NA)
    if (!is.na(pkg)) {
      cat(sprintf("theta: extRemes %.3f | project %.3f (consistency)\n",
                  pkg, own))
      return(pkg)
    }
  }
  own
}
thC <- theta_route(xC, mC$U[, K])
thX <- theta_route(xX, mX$U[, K])
cat(sprintf("theta: NTHMB %.3f (mean run %.1f d), BRIDG %.3f (%.1f d)\n",
            thC, 1 / thC, thX, 1 / thX))
```

# Stage 4: return levels (v8 \S6, eqs. 10--13, pp. 14--15)

```{r rl}
rl_root <- function(Yr, Ts, U, sig, xi, theta) {
  idx <- which(years == Yr)
  u <- U[idx, K]; s <- sig[idx]
  logPr <- function(x, th) {
    z <- pmax(1 + xi * (x - u) / s, 0)
    p <- (1 - params$tau_star) * ifelse(x > u, z^(-1 / xi), 1)
    th * sum(log(pmax(1 - p, 1e-12)))
  }
  sapply(Ts, function(T)
    tryCatch(uniroot(function(x) logPr(x, theta) - log(1 - 1 / T),
                     c(max(u) * 1.0001, max(u) + 80 * max(s)))$root,
             error = function(e) NA))
}
Ygrid <- seq(min(years) + 2, max(years) - 2, by = 5)
Ts <- c(10, 50, 100)
rlC <- t(sapply(Ygrid, rl_root, Ts = Ts, U = mC$U, sig = gC$sig,
                xi = gC$xi, theta = thC))
rlX <- t(sapply(Ygrid, rl_root, Ts = Ts, U = mX$U, sig = gX$sig,
                xi = gX$xi, theta = thX))

## Route A vs Route B at theta = 1: the machinery check (v8 Sec 6, p. 15)
Yr0 <- Ygrid[which.min(abs(Ygrid - median(Ygrid)))]
idx0 <- which(years == Yr0)
u0 <- mX$U[idx0, K]; s0 <- gX$sig[idx0]
n_sim <- 4000
sim_max <- replicate(n_sim, {
  exc <- runif(length(idx0)) < (1 - params$tau_star)
  if (!any(exc)) return(max(u0))
  ex <- which(exc)
  max(u0[ex] + s0[ex] / gX$xi * (runif(length(ex))^(-gX$xi) - 1))
})
routeA_1 <- rl_root(Yr0, Ts, mX$U, gX$sig, gX$xi, theta = 1)
routeB_1 <- quantile(sim_max, 1 - 1 / Ts)
knitr::kable(data.frame(T = Ts, routeA = round(routeA_1, 1),
                        routeB = round(routeB_1, 1),
                        rel_gap = signif(abs(routeA_1 - routeB_1) / routeA_1, 2)),
             caption = sprintf("Route A vs Route B at theta = 1, year %d,
             BRIDG (%d simulated years). Small gaps validate the chaining
             machinery before the theta correction is trusted.", Yr0, n_sim))

op <- par(mfrow = c(1, 2), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
cols <- c("grey55", "steelblue3", "firebrick3")
for (nm in list(list(rlC, "NTHMB"), list(rlX, "BRIDG"))) {
  matplot(Ygrid, nm[[1]], type = "l", lty = 1, lwd = 2, col = cols,
          xlab = "reference year", ylab = "flow",
          main = paste0(nm[[2]], ": x_T(Y), theta-corrected"))
  legend("topleft", paste0("T = ", Ts), col = cols, lwd = 2, bty = "n",
         cex = 0.85)
}
par(op)
```

**Reading the figure.** The x-axis is the reference year $Y$; the y-axis
the flow level $x_T(Y)$ a year with year-$Y$'s climate exceeds with
probability $1/T$ (v8 eq. 10, p. 14 --- a probability statement about a
reference year, not a forecast). One curve per return period, corrected by
$\theta$ for clustering (v8 eq. 13, p. 15). Rising = intensification;
falling is physically possible on snowmelt rivers; the horizontal reading
--- the level that was a 100-year event early becoming a 50-year event
late --- is the paper's most quotable sentence, once the bootstrap bands
of the uncertainty section are around the curves.

# Stage 5: the three conditional functionals (v8 \S6, eq. 14, p. 16)

The professor's three summaries, read off one generated cloud per reference
year, all at the display day of year: the conditional mean, the conditional
quantiles, and the band probability --- demonstrated twice: $P_{co}$
(BRIDG above its own day's $u(t;\tau^{*})$ while NTHMB is extreme) and
$P_{band}$ (bounds at the day's marginal `r params$band_lo` and
`r params$band_hi` quantiles). Bounds and conditioning level are YAML
parameters. On the Laplace scale the band bounds are flat numbers, so the
probabilities need no margin inversion; the mean does, through the day's
$F^{-1}$.

```{r condsummaries}
cond_summaries <- function(level, tgrid_yr, cv_full, Ztab, ladderX, gpX) {
  y_lo <- qlap(params$band_lo); y_hi <- qlap(params$band_hi)
  t(sapply(tgrid_yr, function(Yr) {
    pdE <- at_day(cv_full, Yr)
    yc <- qlap(level + (1 - level) * runif(params$n_mc))
    z <- sample(Ztab, params$n_mc, replace = TRUE)
    yx <- pdE$a * yc + yc^pdE$b * (pdE$m + pdE$s * z)
    px <- pmin(pmax(plap(yx), 1 / (4 * N)), 1 - 1e-9)
    flows <- q_inv(px, ladderX, gpX, pdE$idx)
    c(mean = mean(flows),
      q05 = unname(quantile(flows, 0.05)),
      q50 = unname(quantile(flows, 0.50)),
      q95 = unname(quantile(flows, 0.95)),
      P_co = mean(yx > y_star),
      P_band = mean(yx >= y_lo & yx <= y_hi))
  }))
}
csum <- cond_summaries(params$cond_level, Ygrid, cs_full, Ztable, mX, gX)

op <- par(mfrow = c(1, 2), mar = c(3.5, 3.5, 2.5, 1), mgp = c(2.2, 0.7, 0))
plot(Ygrid, csum[, "mean"], type = "l", lwd = 2, col = "steelblue4",
     ylim = range(csum[, c("q05", "q95")]), xlab = "reference year",
     ylab = "BRIDG flow",
     main = sprintf("E and quantiles of BRIDG | NTHMB at %.2f, doy %d",
                    params$cond_level, doy_star), cex.main = 0.9)
polygon(c(Ygrid, rev(Ygrid)), c(csum[, "q05"], rev(csum[, "q95"])),
        col = adjustcolor("steelblue4", 0.18), border = NA)
lines(Ygrid, csum[, "mean"], lwd = 2, col = "steelblue4")
lines(Ygrid, csum[, "q50"], lwd = 1.5, lty = 2, col = "steelblue4")
legend("topleft", c("mean", "median", "5-95% band"), bty = "n", cex = 0.8,
       col = "steelblue4", lty = c(1, 2, NA), lwd = c(2, 1.5, NA),
       fill = c(NA, NA, adjustcolor("steelblue4", 0.18)),
       border = c(NA, NA, NA))
plot(Ygrid, csum[, "P_co"], type = "b", pch = 16, lwd = 2,
     col = "firebrick3", ylim = c(0, max(csum[, "P_co"]) * 1.15),
     xlab = "reference year", ylab = "probability",
     main = "P_co: BRIDG above its own threshold | NTHMB extreme",
     cex.main = 0.9)
lines(Ygrid, csum[, "P_band"], type = "b", pch = 17, lwd = 2,
      col = "darkorange3")
legend("topleft", c("P_co", sprintf("P_band (%.3g-%.3g)", params$band_lo,
                                    params$band_hi)),
       col = c("firebrick3", "darkorange3"), pch = c(16, 17), lwd = 2,
       bty = "n", cex = 0.8)
par(op)

yr3 <- Ygrid[c(1, ceiling(length(Ygrid) / 2), length(Ygrid))]
knitr::kable(round(csum[Ygrid %in% yr3, ], 3),
             caption = sprintf("The three-year table (years %s, doy %d):
             the three functionals at conditioning level %.2f.",
             paste(yr3, collapse = ", "), doy_star, params$cond_level))
```

**Reading the figure and the table.** Left: x-axis reference year, y-axis
BRIDG flow. The solid line is the conditional mean given NTHMB extreme at
the stated level *on the freshet-peak day* --- the day-of-year conditioning
the meeting required, so these are peak-season answers, comparable with
real annual extremes; the dashed line is the conditional median; the band
spans the conditional 5--95\% quantiles. Right: $P_{co}$ (red) is the
co-exceedance probability --- BRIDG above its own extreme threshold while
NTHMB is extreme; $P_{band}$ (orange) the chance BRIDG lands between its
day's marginal `r params$band_lo` and `r params$band_hi` quantiles on those
days. Both rising along the axis is co-flooding intensifying even where
marginal extremes weaken. The table is the same three functionals at the
axis start, middle and end --- the tabular form of "how much worse."

## The HT04 Table-5 discipline: model vs empirical, where the data can check

```{r table5}
eras2 <- list(early = min(years):(min(years) + 29),
              late = (max(years) - 29):max(years))
t5 <- do.call(rbind, lapply(names(eras2), function(en) {
  yrs_e <- eras2[[en]]
  emp_days <- which(keep & years %in% yrs_e)
  emp <- mean(xX[emp_days])
  ## generate at the OBSERVED exceedance days of the era (their own year
  ## and day of year), so the comparison is day-matched, not day-averaged
  gen_days <- sample(emp_days, params$n_mc, replace = TRUE)
  yc <- qlap(params$kappa + (1 - params$kappa) * runif(params$n_mc))
  z <- sample(Ztable, params$n_mc, replace = TRUE)
  a_d <- cs_full$a[gen_days]; b_d <- cs_full$b[gen_days]
  m_d <- cs_full$m[gen_days]; s_d <- cs_full$s[gen_days]
  yx <- a_d * yc + yc^b_d * (m_d + s_d * z)
  px <- pmin(pmax(plap(yx), 1 / (4 * N)), 1 - 1e-9)
  gen <- mean(sapply(seq_len(params$n_mc), function(i)
    q_inv(px[i], mX, gX, gen_days[i])))
  data.frame(era = en, empirical = round(emp, 1), model = round(gen, 1),
             rel_gap = round(abs(emp - gen) / emp, 3))
}))
knitr::kable(t5, row.names = FALSE,
             caption = "Model vs empirical conditional mean of BRIDG flow
             given NTHMB above the dependence threshold, per era, generated
             at the observed exceedance days themselves (day-matched).
             Agreement here is the licence for the higher-level
             extrapolations above (HT04 Table 5 discipline).")
```

# Uncertainty: the block bootstrap (v8 \S6, p. 15)

```{r boot}
boot_once <- function() {
  yrs <- unique(years)
  keep_y <- years %in% sample(yrs, replace = TRUE)
  gX_b <- pen_rq_seas(xX[keep_y], Xseas[keep_y, ], params$tau_star,
                      ifelse(is.na(mX$lam_u), 100, mX$lam_u))
  uX_b <- as.vector(Xseas %*% gX_b)
  e <- xX > uX_b & keep_y
  gpX_b <- fit_gp_basis((xX - uX_b)[e], Xseas[e, , drop = FALSE],
                        gX$lam_s, Dtr)
  th_b <- ferro_segers(which(xX > uX_b))
  kb <- YC > v_dep & keep_y
  hb <- fit_ht_basis(YC[kb], YX[kb], cbind(1, tt[kb]),
                     Bchk = cbind(1, tdec))
  sigX_b <- exp(as.vector(Xseas %*% gpX_b$gam))
  U_b <- mX$U; U_b[, K] <- uX_b
  list(rl100 = sapply(Ygrid, function(Yr)
         rl_root(Yr, 100, U = U_b, sig = sigX_b, xi = gpX_b$xi,
                 theta = th_b)),
       a_ends = 2 * plogis(hb$par[1] + hb$par[2] * c(0, 1)) - 1)
}
boots <- replicate(params$B_boot, boot_once(), simplify = FALSE)
rl_mat <- sapply(boots, `[[`, "rl100")
a_mat <- sapply(boots, `[[`, "a_ends")
bandr <- apply(rl_mat, 1, quantile, c(0.025, 0.975), na.rm = TRUE)
plot(Ygrid, rlX[, 3], type = "n",
     ylim = range(bandr, rlX[, 3], na.rm = TRUE),
     xlab = "reference year", ylab = "flow",
     main = sprintf("BRIDG: x_100(Y), %d-replicate year-block bootstrap
     (project fitters inside the loop; selections join at B = 2000)",
                    params$B_boot))
polygon(c(Ygrid, rev(Ygrid)), c(bandr[1, ], rev(bandr[2, ])),
        col = adjustcolor("firebrick3", 0.2), border = NA)
lines(Ygrid, rlX[, 3], lwd = 2, col = "firebrick3")

a_ci <- apply(a_mat, 1, quantile, c(0.025, 0.975))
a_lin_ends <- 2 * plogis(h1$par[1] + h1$par[2] * c(0, 1)) - 1
knitr::kable(data.frame(endpoint = c("axis start", "axis end"),
  a_hat = round(a_lin_ends, 3),
  lo = round(a_ci[1, ], 3), hi = round(a_ci[2, ], 3)),
  caption = "The honest move-one test at the dependence stage (v8 Sec 4.6,
  p. 10): year-block bootstrap intervals for a(t) at the axis endpoints,
  Laplace scale.")
```

**Reading the figure and the table.** The red curve is the production
100-year level for BRIDG; the band its bootstrap 95\% interval --- years
resampled with their labels kept, threshold, GP, $\theta$ and the linear
dependence model refitted per replicate, so the band measures estimation
uncertainty on the same axis (v8 \S6, p. 15). A band at the late axis that
excludes the early-axis level is the uncertainty-aware "the risk changed."
The table gives the bootstrap answer for the dependence trend: the interval
at the axis end excluding the interval at the start is what certifies the
working-likelihood LRT's verdict (v8 \S4.6, p. 10). At $B = `r params$B_boot`$
these are knit-scale indications; the paper-scale run is $B = 2000$ with
every selection inside the loop.

# The RCP 8.5 arm (v8 \S6, p. 16)

The scenario comparison --- the display contrast the meeting settled on ---
needs the RCP 8.5 series in the same schema as this file (station, date,
sf, bf; 1969--2099). Once that export exists, this report's pipeline runs
on it unchanged (the axis is read from the file), and the
scenario-contrast versions of the pseudosample display and the functionals
replace the within-scenario year contrast. Blocked on the data export;
nothing else here waits on it.

# Standing items, and the Friday list

**Known defect (blocking, Aug-22 run):** the Keef constraint's hard wall
stalled the BFGS optimiser at the dependence stage -- all model sizes
returned the identical objective, the trend statistic was exactly 0.00, and
the constraint-off refit found a far lower objective with a rising a(t). No
dependence-stage number (including the functionals) is quotable until the
smooth-penalty repair is in; the full diagnosis, evidence and repair recipe
are in walkthrough_v4 (same folder), Section "the defect".

Cleaning: the forensics tables above are evidence, not decisions ---
which values or months to exclude, and how to document it, is decided with
the professor; 1968 is already out. The three problems (data, threshold,
excesses + PIT) map to Sections 2, 3 and 4--5. For Friday: confirm the
identity of "the three problems"; the display scale (standard vs original,
Section 7 draws both); whether the year contrast on the original-scale
panel is wanted alongside the scenario contrast; and the seasonal-threshold
consequence flagged in the preamble --- with margins that bend within the
year, the dependence stage conditions on season-relative extremes, and that
reading should be agreed explicitly. Laplace is the working scale
throughout (v8 \S3); the Gumbel refit remains available as the sensitivity
appendix v8 \S7 (p. 17) names. Reproducibility: seed 20260818 once at the
top; knit-scale settings are the YAML parameters.

```{r session}
sessionInfo()
```
