---
title: "Baseflow Full EDA Report (1945-2012)"
subtitle: "Temporal Aggregation and Spatial Coherence Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = TRUE)
library(terra)
library(ggplot2)
library(dplyr)
library(viridis)
library(leaflet)
library(htmltools)

# -------------------------------------------------------
# DATA LOADING
# -------------------------------------------------------
# Ensure these files exist in your directory
yearly_max <- rast("yearly_max_1945_2012.tif")
nc_path <- "Gridded 1945-2012 Baseflow.nc"
bf_stack <- rast(nc_path)["BASEFLOW"]

# Pre-calc date information
all_dates <- time(bf_stack)

```

## 1. Temporal Drift: Is the Region Getting Wetter?

We analyze if the maximum baseflow intensity has increased over the 68-year period.

```{r trend_analysis}
# 1. Calculate the regional average of the yearly maximums
# Using 'yearly_max' is efficient here
regional_annual_max <- global(yearly_max, "mean", na.rm = TRUE)

# 2. Create a dataframe for plotting
df_trend <- data.frame(
  Year = 1945:2012,
  Max_Baseflow = regional_annual_max$mean
)

# 3. Plot with a trend line (Loess smoothing)
ggplot(df_trend, aes(x = Year, y = Max_Baseflow)) +
  geom_line(color = "steelblue", alpha = 0.5) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "loess", color = "darkred", fill = "pink") +
  theme_minimal() +
  labs(title = "Average Annual Maximum Baseflow (1945-2012)",
       subtitle = "Aggregated across the Vancouver/PNW Region",
       y = "Baseflow (mm/day)",
       x = "Year")

```

---

## 2. Spatial Coherence: The Event of Jan 15, 1997

We test the hypothesis that **January 15, 1997** was a regionally coherent event where >65% of the region exceeded the 90th percentile flows.

```{r spatial_coherence}
# -------------------------------------------------------
# 1. Define the Event
# -------------------------------------------------------
target_date_str <- "1997-01-15"
event_idx <- which(as.Date(all_dates) == as.Date(target_date_str))

if(length(event_idx) > 0) {
  
  # -------------------------------------------------------
  # 2. Calculate the 90th Percentile Threshold (The "High Flow" Mark)
  # Note: Calculating quantile on the full daily stack is heavy. 
  # Ideally, do this once and save it. Here we compute it on the fly.
  # -------------------------------------------------------
  
  # To save time for this report, we check if the threshold file exists, otherwise calculate
  if(file.exists("bf_quantile_90.tif")) {
    bf_q90 <- rast("bf_quantile_90.tif")
  } else {
    # WARNING: This app function takes time on large stacks!
    # If too slow, use a smaller sample or load pre-computed stats.
    bf_q90 <- quantile(bf_stack, probs = 0.90, na.rm=TRUE)
    # writeRaster(bf_q90, "bf_quantile_90.tif") # Optional: save for later
  }

  # -------------------------------------------------------
  # 3. Compare Event Day to Threshold
  # -------------------------------------------------------
  event_layer <- bf_stack[[event_idx]]
  
  # Create a Binary Map: 1 if Event > 90th Percentile, 0 otherwise
  exceedance_map <- event_layer > bf_q90
  
  # Calculate Percentage of Area
  global_exceedance <- global(exceedance_map, "mean", na.rm=TRUE) # Mean of 0/1 is the %
  pct_area <- round(global_exceedance$mean * 100, 2)
  
  print(paste("On", target_date_str, pct_area, "% of the region exceeded the 90th percentile baseflow."))

  # -------------------------------------------------------
  # 4. Visualization
  # -------------------------------------------------------
  # Plot Side-by-Side
  par(mfrow = c(1, 2))
  
  # Plot 1: Raw Baseflow
  plot(event_layer, 
       main = paste("Raw Baseflow:", target_date_str), 
       col = viridis(100), 
       range = c(0, max(values(event_layer), na.rm=T)))
  
  # Plot 2: Spatial Coherence (Binary)
  plot(exceedance_map, 
       main = "Exceedance of 90th Percentile", 
       col = c("lightgrey", "red"), 
       legend = FALSE)
  legend("bottomright", legend = c("Normal", "> 90th Pctl"), fill = c("lightgrey", "red"))
  
  par(mfrow = c(1, 1)) # Reset
  
} else {
  print("Event date not found in dataset.")
}

```

---

## 3. Interactive Maps (Leaflet)

This map visualizes the specific event date.

```{r interactive_map, eval=TRUE}
# NOTE: eval=TRUE is required for this to show up!

target_date_viz <- "1997-01-15"
viz_idx <- which(as.Date(all_dates) == as.Date(target_date_viz))

if (length(viz_idx) > 0) {
  
  layer_to_plot <- bf_stack[[viz_idx]]
  
  # Define palette
  vals <- values(layer_to_plot, mat=FALSE)
  # Cap the legend at 98th percentile to avoid outliers skewing colors
  max_val <- quantile(vals, 0.98, na.rm=TRUE) 
  
  pal <- colorNumeric(c("#FFFFCC", "#41B6C4", "#0C2C84"), 
                      domain = c(0, max_val),
                      na.color = "transparent")

  leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    addRasterImage(layer_to_plot, 
                   colors = pal, 
                   opacity = 0.8,
                   project = TRUE, # IMPORTANT: Must project to Web Mercator
                   maxBytes = 30 * 1024 * 1024) %>% # Increased size limit
    addLegend(pal = pal, 
              values = c(0, max_val),
              title = paste("Baseflow (mm)<br>", target_date_viz))
} else {
  print("Date not available for mapping.")
}

```

---

## 4. Point Extraction (Location Specific)

Extracting time series for a specific coordinate (e.g., approximate location of North Vancouver).

```{r point_extraction}
# 1. Define specific coordinates (Longitude, Latitude)
# Example: North Vancouver / Capilano area
target_lon <- -123.10
target_lat <- 49.35
target_vect <- vect(cbind(target_lon, target_lat), crs="+proj=longlat +datum=WGS84")

# 2. Extract
# Ensure coords are in same CRS as raster. If raster is projected, project the point.
if (crs(bf_stack) != crs(target_vect)) {
  target_vect <- project(target_vect, crs(bf_stack))
}

point_vals <- terra::extract(bf_stack, target_vect, ID=FALSE)

# 3. Check if we hit NA (ocean). If so, search nearby.
if (all(is.na(point_vals))) {
  print("Selected point is NA (likely ocean). finding nearest land pixel...")
  # Simple workaround: take nearest non-NA cell
  cells <- nearby(bf_stack[[1]], cells=cellFromXY(bf_stack, xy=crds(target_vect)), distance=5000)
  # (Requires more complex logic for nearest neighbor, usually manual adjustment is best)
} 

# 4. Plotting
# Transpose result to a vector
ts_data <- as.numeric(t(point_vals))

df_point <- data.frame(
  Date = all_dates,
  Baseflow = ts_data
) %>%
  filter(!is.na(Baseflow))

# Highlight the 1997 event
event_val <- df_point %>% filter(Date == as.Date("1997-01-15"))

ggplot(df_point, aes(x = Date, y = Baseflow)) +
  geom_line(color = "darkgreen", linewidth = 0.5) +
  # Add red dot for the event
  geom_point(data = event_val, aes(x=Date, y=Baseflow), color="red", size=3) +
  geom_text(data = event_val, aes(label="Jan 1997 Event", y=Baseflow), vjust=-1, color="red") +
  theme_light() +
  labs(title = paste("Local Baseflow Series"), 
       subtitle = paste("Location: ", target_lat, "N, ", target_lon, "W"),
       y = "Baseflow (mm)")

```

## 5. Record Breaker Map

Years where the historic maximum occurred.

```{r record_breaker}
# Find the index (year) of the maximum value for every pixel
year_of_max_idx <- app(yearly_max, which.max)

# Adjust index to Year
historic_max_years <- year_of_max_idx + 1944

# Plot
plot(historic_max_years, 
     main = "Year of Historic Maximum Baseflow",
     col = terra::map.pal("viridis", 10),
     type = "continuous")

```

```

```